---
title: "03 — Climate"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = FALSE, message = FALSE, warning = FALSE)
options(scipen = 10000, digits = 12)
set.seed(711)
```

```{r load-packages, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(vegan)
library(caret)
library(corrplot)
library(dismo)
library(ggpubr)
library(ggrepel)
library(sf)
library(rnaturalearth)
library(ggnewscale)
```

```{r load-globals, message=FALSE, warning=FALSE}
source(here::here("R/colors_themes.R"))
source(here::here("R/rf_functions.R"))
source(here::here("R/utils.R"))
```

```{r load-rds, message=FALSE, warning=FALSE}
metadata <- readRDS(here::here("results/rds/metadata_litter.rds"))
pca_litt_df_md <- readRDS(here::here("results/rds/pca_litt_df_md.rds"))
PCAloads_litt_md <- readRDS(here::here("results/rds/PCAloads_litt_md.rds"))
reserve_codenames <- readRDS(here::here("results/rds/reserve_codenames.rds"))
```

##### Site map

Map of the 24 UC Natural Reserve System sites, coloured by ecosystem type. Note that in the metadata `plot_cordsX` holds latitude and `plot_cordsY` holds longitude.

```{r site-map, message=FALSE, warning=FALSE, fig.width=9, fig.height=7, results='hold', echo=FALSE}
# California state boundary (1:50m Natural Earth), cached to avoid re-downloading
california <- cached(here::here("cache/california_sf.rds"), {
  rnaturalearth::ne_download(scale = 50, type = "admin_1_states_provinces_lakes",
                             category = "cultural", returnclass = "sf") |>
    dplyr::filter(name == "California")
})

# One point per reserve (plot_cordsX = latitude, plot_cordsY = longitude);
# fix the "Connifer Forest" spelling so it matches ecosystem_colors, and attach short codes
sites_sf <- metadata |>
  dplyr::distinct(Reserve, .keep_all = TRUE) |>
  dplyr::select(Reserve, Ecosystem, lat = plot_cordsX, lon = plot_cordsY) |>
  dplyr::mutate(Ecosystem = dplyr::recode(Ecosystem, "Connifer Forest" = "Conifer Forest")) |>
  dplyr::left_join(reserve_codenames |> dplyr::select(ReserveName, CodeName),
                   by = c("Reserve" = "ReserveName")) |>
  dplyr::arrange(CodeName) |>
  dplyr::mutate(reserve_key = factor(paste0(CodeName, " = ", Reserve),
                                     levels = paste0(CodeName, " = ", Reserve))) |>
  sf::st_as_sf(coords = c("lon", "lat"), crs = 4326)

# colours for the reserve-key legend = each reserve's ecosystem colour
reserve_key_cols <- setNames(ecosystem_colors[sites_sf$Ecosystem], sites_sf$reserve_key)

site_map <- ggplot() +
  geom_sf(data = california, fill = "grey97", colour = "grey55", linewidth = 0.3) +
  geom_sf(data = sites_sf, aes(colour = Ecosystem), size = 2, alpha = 0.9) +
  scale_colour_manual(values = ecosystem_colors, name = "Ecosystem") +
  # short codes on the map
  geom_text_repel(data = sites_sf, aes(geometry = geometry, label = CodeName),
                  stat = "sf_coordinates", size = 2.6, fontface = "bold",
                  max.overlaps = Inf, min.segment.length = 0,
                  segment.size = 0.2, box.padding = 0.3) +
  coord_sf(expand = FALSE) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid = element_line(colour = "grey92"),
        legend.text = element_text(size = 9),
        legend.key.size = unit(0.35, "cm"),
        legend.position = "inside",
        legend.position.inside = c(0.85, 0.8))
site_map

ggsave(here::here("plots/site_map_california.png"), site_map,
       width = 90, height = 90, dpi = 300, bg = "white", units = "mm")
```

##### Climate data analysis

I downloaded the monthly normals (averages over 30 years 1991-2020) of max and min temperature and precipitation data from PRISM (U. of Oregon). I obtained values for each of the coordinates for the Grass and Dominant Vegetation plots.

```{r climate-data, message=FALSE, warning=FALSE, results='hold', echo=FALSE}
prism_data <- read.csv(here::here("PRISM_ppt_tmin_tmean_tmax_30yr_normal_800m_monthly_normals.csv"),
                       header = T)
prism_data$Name <- as.factor(prism_data$Name)
prism_data <- prism_data %>% filter(!is.na(Longitude))

bioclim_vars <- data.frame()
for (i in 1:length(unique(prism_data$Name))) {
  filter_temp  <- prism_data %>% filter(Name == unique(prism_data$Name)[i] & Date != "Annual")
  biovars_temp <- biovars(filter_temp$ppt..mm., filter_temp$tmin..degrees.C., filter_temp$tmax..degrees.C.)
  tmp          <- data.frame(reserve_litter = unique(prism_data$Name)[i], biovars_temp)
  bioclim_vars <- rbind(bioclim_vars, tmp)
  rm(tmp)
}

metadata <- metadata %>%
  mutate(reserve_litter = paste(reserve_code, Litter_type, sep = "-")) %>%
  merge(., bioclim_vars, by = "reserve_litter")

bioclim_vars_md <- metadata |> distinct(reserve_litter, .keep_all = T)

bioclim_names <- read.table(here::here("data/bioclim_names.txt"), sep = "\t")

bioclim_labels <- c(
  bio1 = "MAT", bio2 = "T_Diurn", bio3 = "T_Iso", bio4 = "T_Seas",
  bio5 = "T_MaxWarmMo", bio6 = "T_MinColdMo", bio7 = "T_Range",
  bio8 = "T_WetQ", bio9 = "T_DryQ", bio10 = "T_WarmQ", bio11 = "T_ColdQ",
  bio12 = "MAP", bio13 = "P_WetMo", bio14 = "P_DryMo", bio15 = "P_Seas",
  bio16 = "P_WetQ", bio17 = "P_DryQ", bio18 = "P_WarmQ", bio19 = "P_ColdQ"
)

colnames(bioclim_vars_md)[22:40] <- bioclim_labels

# Find correlated variables
nzv <- nearZeroVar(bioclim_vars_md[, c(22:40)])
nzv
cor.matrix     <- cor(bioclim_vars_md[, c(22:40)], use = "na.or.complete")
corrplot(cor.matrix, order = "hclust", method = "color", type = "lower",
         tl.cex = 0.7, tl.col = rgb(0, 0, 0), addCoef.col = T)
highlyCorrelated <- findCorrelation(cor.matrix, cutoff = 0.85, names = T, exact = T, verbose = T)
print(highlyCorrelated)
highlyCorrelated <- c(highlyCorrelated[highlyCorrelated != "MAP"], "P_ColdQ")


```

###### Climate by plot

```{r climate-by-plot, message=FALSE, warning=FALSE, results='hold', echo=FALSE, fig.width=30, fig.height=25}
bioclim_md <- setNames(metadata[22:40], bioclim_labels)
metadata   <- cbind(metadata[, -c(22:40)], bioclim_md)
rownames(metadata) <- metadata$SampleID

pca <- prcomp(metadata[, c(22:40)] |> distinct(.keep_all = T), center = T, scale. = T)
pcadf <- data.frame(pca$x) |>
  dplyr::select(PC1, PC2) %>%
  merge(., metadata, by = "row.names")

climate_PERMANOVA <- adonis2(
  dist(scale(metadata[, c(22:40)] |> distinct(.keep_all = T))) ~ Ecosystem,
  data = metadata[, c(4, 22:40)] |> distinct(.keep_all = T) |> dplyr::select(Ecosystem),
  permutations = 999)
climate_PERMANOVA

PCAloads    <- data.frame(Variables = rownames(pca$rotation), pca$rotation)
summary(pca)

pca_sq_loadings  <- pca$rotation^2
pca_contributions <- sweep(pca_sq_loadings, 2, colSums(pca_sq_loadings), FUN = "/") * 100
print(pca_contributions)

pca_plot <- ggplot(pcadf, aes(PC1, PC2)) +
  geom_point(data = pcadf, aes(color = Ecosystem), size = 4, alpha = 0.75) +
  scale_color_manual(values = ecosystem_colors, name = "Ecosystem") +
  geom_segment(data = PCAloads,
               aes(x = 0, xend = PC1 * 10, y = 0, yend = PC2 * 10),
               colour = "grey", inherit.aes = FALSE) +
  ggrepel::geom_text_repel(data = PCAloads,
                            aes(x = PC1 * 10.7, y = PC2 * 10.7,
                                label = gsub("_", " ", Variables)),
                            size = 4,
                            colour = c(rep("#990000", 11), rep("darkblue", 8)),
                            alpha = 0.95, fontface = "bold") +
  xlab("PC1 (44.55%)") + ylab("PC2 (31.42%)") +
  guides(color = guide_legend(title.position = "top"), shape = "none") +
  apatheme +
  theme(legend.position = c(0.9, 0.85),
        legend.background = element_rect(fill = alpha("white", 0.6)),
        legend.key.size = unit(0.3, "cm"), legend.spacing.y = unit(0.15, "cm"),
        legend.margin = ggplot2::margin(2, 2, 2, 2),
        legend.text = element_text(size = 10),
        legend.title = element_text(size =11))
pca_plot

pca_contributions <- as.data.frame(pca_contributions) |>
  rownames_to_column(var = "Variable") |>
  pivot_longer(cols = c(2:20), names_to = "PC", values_to = "Percent") |>
  mutate(Variable = factor(Variable, levels = colnames(metadata[22:40])))

pca_contributions_plot <- pca_contributions |>
  filter(PC %in% c("PC1", "PC2")) |>
  ggplot(aes(x = PC, y = Percent, fill = Variable)) +
  geom_col() +
  scale_fill_manual(values = clim_pallete, labels = sub("_", " ", unique(pca_contributions$Variable))) +
  labs(y = "% of variance contribution") +
  apatheme + theme(axis.title.x = element_blank())
pca_contributions_plot

clim_plot_final <- ggarrange(pca_plot, pca_contributions_plot,
                             nrow = 1, widths = c(1, 0.5),
                             labels = c("1A.1", "1A.2"), hjust = 0, vjust = 1)
clim_plot_final

ggsave(here::here("plots/PCAs/clim_pca_ecosystem.png"),
       width = 300, height = 150, units = "mm", dpi = 300, bg = "white")
```

###### Printing PCA plots

```{r print-pca-plots}
ggarrange(pca_plot, pca_plot_litt_md, ncol = 2, common.legend = F, labels = "AUTO")
ggsave(here::here("plots/PCAs/clim&litter_PCA.png"),
       width = 280, height = 140, units = "mm", dpi = 300, bg = "white")

ggarrange(pca_contributions_plot, pca_litt_contributions_plot_md,
          ncol = 2, common.legend = F, labels = "AUTO")
ggsave(here::here("plots/PCAs/clim&litter_PCA_contributions.png"),
       width = 240, height = 130, units = "mm", dpi = 300, bg = "white")
```

```{r save-rds, include=FALSE}
saveRDS(metadata, here::here("results/rds/metadata_climate.rds"))
saveRDS(pcadf, here::here("results/rds/pcadf_clim.rds"))
saveRDS(PCAloads, here::here("results/rds/PCAloads_clim.rds"))
saveRDS(highlyCorrelated, here::here("results/rds/highlyCorrelated.rds"))
saveRDS(bioclim_labels, here::here("results/rds/bioclim_labels.rds"))
```
