---
title: "10 — Partial Dependence Plots"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = FALSE, message = FALSE, warning = FALSE)
options(scipen = 10000, digits = 12)
set.seed(711)
```

```{r load-packages, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(iml)
library(randomForest)
library(rfPermute)
library(A3)
library(caret)
library(ggpubr)
library(scales)
library(ggh4x)
library(reshape2)
```

```{r load-globals, message=FALSE, warning=FALSE}
source(here::here("R/colors_themes.R"))
source(here::here("R/rf_functions.R"))
source(here::here("R/utils.R"))
```

```{r load-rds, message=FALSE, warning=FALSE}
clades_Ab_md       <- readRDS(here::here("results/rds/clades_Ab_md.rds"))
subclades_Ab_md    <- readRDS(here::here("results/rds/subclades_Ab_md.rds"))
clim_vars_PC       <- readRDS(here::here("results/rds/clim_vars_PC.rds"))
litter_vars        <- readRDS(here::here("results/rds/litter_vars.rds"))
subclade_colors_bp <- readRDS(here::here("results/rds/subclade_colors_bp.rds"))
subclade_colros    <- readRDS(here::here("results/rds/subclade_colros.rds"))
clade_colors       <- readRDS(here::here("results/rds/clade_colors.rds"))
correct_order      <- readRDS(here::here("results/rds/correct_order.rds"))
count_summary      <- readRDS(here::here("results/rds/count_summary.rds"))
pcadf_expanded     <- readRDS(here::here("results/rds/pcadf_expanded.rds"))
```

#### Partial dependance plots

#### Partial dependance plots

##### PDP clade level with Random Forest model

```{r pdp-clades-rf, message=FALSE, warning=FALSE, results='hold', echo=FALSE, fig.width=25, fig.height=20,}
library(iml)
library(randomForest)
set.seed(711)

# Set up an output directory to save the plots
output_dir <- here::here("plots/plots_pdp")
#dir.create(output_dir, showWarnings = FALSE)

## add PCs from PCAs for litter chem, climate and all variables


clim_vars <- #gsub("\\.", " ",
                  c("MAT",
              "MAP",
              "P.Seas",
              "P.WarmQ",
              "T.Iso",
              "T.Diurn",
              "T.MinColdMo",
              "T.DryQ",
              "P.DryMo")#)

clim_vars_PC <- c("PC1.clim",
              "PC2.clim")

litter_vars <- c("pH", "crude_protein", "cellulose", "hemicellulose", "lignin")

litter_vars_PC <- c("PC1.litt", "PC2.litt")

#all_vars_PC <- c("PC1.all.vars", "PC2.all.vars")

### let's prepare also a pallete with spaces on the names of the variables

allvars_pallete_5 <- allvars_pallete

names(allvars_pallete_5) <- gsub("_", " ", names(allvars_pallete_5))


# Replace spaces with underscores in column names
colnames(clades_Ab_md) <- gsub(" ", ".", colnames(clades_Ab_md))


## climate PCs and litter variables

#initialize empty data frames to store results
imp_results <- data.frame()
effs_results <- data.frame()

#Set a loop to go over each clade
for (clade in unique(clades_Ab_md$Clade)) {
  #Set a loop to go over each type of litter
  #for (litter in unique(clades_Ab_md$Litter_type)) {
    #subset for each clade and each type of litter
    pdp_temp <- clades_Ab_md %>%
      filter(Clade == clade# & Litter_type == litter
                                ) %>%
      ungroup() %>%
      dplyr::select(-c(Row.names, reserve_code, Litter_type, Clade)) |>
      dplyr::select(all_of(c(litter_vars, clim_vars_PC)), Abundance)
    #create a random forest for each subseted dataset
    rf <- randomForest(Abundance ~., data = pdp_temp, ntree = 1000)
    #get the variables for Abundace prediction
    X <- pdp_temp[!(names(pdp_temp) %in% "Abundance")]
    #set model variables for prediction of Abundance
    predictor <- Predictor$new(rf, data = X, y = pdp_temp$Abundance)
    #obtain importance of varibles in the model
    imp <- FeatureImp$new(predictor, loss = "rmse", n.repetitions = 100, compare = "difference")
    
    #store importance results into a data frame and add clade and litter type information
    imp_temp <-imp$results
    imp_temp$Clade <- clade
    #imp_temp$Litter_type <- litter
    
    imp_results <- rbind(imp_results, imp_temp)
    
    #calculate predictions
    effs <- FeatureEffects$new(predictor, grid.size = 20, method = "pdp")
    effs_temp <- do.call(rbind, effs$results)
    effs_temp$Clade <- clade
    #effs_temp$Litter_type <- litter
   
    effs_results <- rbind(effs_results, effs_temp)
    
    rug <- X
    #object_name <- paste("rug", litter, sep = "_")
    #assign(object_name, X)
  #}
}


clade_results <- cached(here::here("cache/clade_results.rds"), {
  res_list <- list()
  for (clade in unique(clades_Ab_md$Clade)) {
    message("Processing clade: ", clade)
    df_cl <- clades_Ab_md %>%
      filter(Clade == clade) %>%
      dplyr::select(all_of(c("Abundance", litter_vars, clim_vars_PC))) %>%
      na.omit()
    res <- run_rf_analysis(
      df           = df_cl,
      response_var = "Abundance",
      predictors   = c(litter_vars, clim_vars_PC),
      ntree        = 1000, cv_folds = 10, nperm_oob = 500,
      nrep_vi      = 500, n_boot = 100, grid.size = 20)
    str(res$imp_results_rf)
    res$summary_table$Clade  <- clade
    res$pdp_ci_df$Clade      <- clade
    res$imp_results_rf$Clade <- clade
    res$rf_cv$results$Clade  <- clade
    res_list[[clade]] <- res
  }
  res_list
})

# Now: combine into single tables

# (a) Combined summary of overall metrics:
all_summaries <- bind_rows(
  lapply(clade_results, `[[`, "summary_table")
)

# (b) Combined PDP+CI for all clades:
all_pdp_ci <- bind_rows(
  lapply(clade_results, `[[`, "pdp_ci_df")
)

# (c) Combined iml importances:
all_imp <- bind_rows(
  lapply(clade_results, `[[`, "imp_results_rf")
)

# (d) Combined CV results (if you like):
all_cv <- bind_rows(
  lapply(clade_results, function(x) x$rf_cv$results)
)


# Make a tibble with one row per clade‐feature pair
sig_lookup <- map_dfr(
  .x   = names(clade_results),
  .f   = function(cl) {
    tibble(
      Clade   = cl,
      feature = clade_results[[cl]]$sig_vars_rfp
    )
  }
)


write.table(all_summaries, file = here::here("plots/RF_validation_clades.csv"), quote = F, row.names = F, col.names = T, sep = ",")


all_imp_sig    <- all_imp    %>% semi_join(sig_lookup, by = c("Clade", "feature"))
#all_pdp_ci_sig <- all_pdp_ci %>% semi_join(sig_lookup, by = c("Clade", "feature"))

# And so on…  
# You can now plot or write out any of these combined tables:

print(all_summaries)
print(all_imp)
#print(all_pdp_ci)


# 0) define your clean, recoded, ordered feature vector
feature_levels <- c("PC1","PC2","lignin","cellulose","hemicellulose","crude protein","pH")
feature_levels_imp <- c("PC1","PC2","lignin","cellulose","hemicellulose","crude protein","pH")
# Palette with PC keys renamed to match post-recode feature names
allvars_pallete_imp <- c(
  allvars_pallete_2[!names(allvars_pallete_2) %in% c("PC1 clim", "PC2 clim")],
  "PC1" = unname(allvars_pallete_2["PC1 clim"]),
  "PC2" = unname(allvars_pallete_2["PC2 clim"])
)

# Subclade -> clade groupings, derived from the subclade IDs
# (each subclade is its clade's Roman numeral plus an uppercase suffix, e.g. IIIA -> III)
.all_subclades   <- sort(unique(subclades_Ab_md$Subclade))
.subclade_clade  <- gsub("[^IV]+$", "", .all_subclades)
clade_I_subclades   <- .all_subclades[.subclade_clade == "I"]
clade_II_subclades  <- .all_subclades[.subclade_clade == "II"]
clade_III_subclades <- .all_subclades[.subclade_clade == "III"]
clade_IV_subclades  <- .all_subclades[.subclade_clade == "IV"]
clade_V_subclades   <- .all_subclades[.subclade_clade == "V"]


# 1) Re‐mutate `all_imp_sig` so that its `feature` column is exactly one of those names:
all_imp_sig <- all_imp_sig %>%
  mutate(
    # 1a) Remove dots or underscores and recode the climate PCs
    feature = gsub("[._]", " ", feature),
    feature = dplyr::recode(feature,
      `PC1 clim` = "PC1",
      `PC2 clim` = "PC2"
    ),
    # 1b) Also make sure “crude_protein” → “crude protein” exactly
    feature = if_else(feature == "crude protein" | feature == "crude_protein", 
                      "crude protein", feature)
    )

# 2) compute finite coords *after* recoding + factoring features
pos_df_clade <- all_imp_sig %>%
  group_by(Clade) %>%
  reframe(
    x = max(importance, na.rm = TRUE) * 1.05,
    y = n() + 0.3           # 0.3 units above the topmost point
  )

# 3) Re‐factor `feature` using the clean `feature_levels`
#all_imp_sig <- all_imp_sig %>%
#  mutate(
#    feature = factor(feature, levels = feature_levels)
#  )



# 1) extract per‐clade CV R² & p‐value
annot_df_clade <- all_summaries %>%
  filter(Metric %in% c("10-fold CV R²", "Permuted CV p-value")) %>%
  dplyr::select(Clade, Metric, Value) %>%
  pivot_wider(names_from = Metric, values_from = Value) %>%
  dplyr::rename(
    CV_R2    = `10-fold CV R²`,
    P_val    = `Permuted CV p-value`
  ) %>%
  mutate(
    CV_R2 = as.numeric(CV_R2),
    P_num = as.numeric(sub("^<", "", P_val)),
    expr  = paste0(
      "atop(",
        "CV~R^2==", round(CV_R2, 3), ",",
        "italic(P)<", format.pval(P_num, digits=2, eps = 1e-3),
      ")"
    )
  )



annot_df_clade <- annot_df_clade %>% left_join(pos_df_clade, by="Clade") |>
  dplyr::filter(!Clade %in% c("V", "UND"))



imp_plot_pdp_rf_clades <- all_imp_sig %>%
  mutate(var_type = case_when(feature %in% clim_vars_PC ~ "Climate",
                              feature %in% litter_vars ~ "Litter chemistry",
                              TRUE ~ "All"),
         Clade = factor(Clade, levels = c("I", "II", "III", "IV", "V", "UND"))) |> 
  dplyr::filter(Clade != "UND") |> 
  mutate(feature = factor(feature, levels = feature_levels_imp)) %>%
  ggplot(aes(x = importance, y = feature, color = feature)) +
  #scale_color_manual(values = clade_colors, limits = c("I", "II", "III", "IV", "V", "UND")) +
  scale_color_manual(values = allvars_pallete_imp,
                     labels = c(PC1 = "PC1 clim", PC2 = "PC2 clim"),
                     guide = guide_legend(ncol = 2)) +
  scale_y_discrete(limits = rev(c("PC1", "PC2", "lignin", "cellulose", "hemicellulose", "crude protein", "pH"))) +
  scale_x_continuous(labels = scales::percent_format()) +
  #facet_grid(Clade~., scales = "free_y", space = "free_y") +
  ggh4x::facet_grid2(Clade ~ ., scales = "free_y", space = "free_y",
                       strip = ggh4x::strip_themed(
                         background_y = ggh4x::elem_list_rect(
                           fill = clade_colors[-5] # one color per Clade
                         )
                       )) +
  geom_point(position = position_dodge(width = 0.4), size = 2.5) +
  geom_errorbarh(aes(xmin = importance.05, xmax = importance.95),
                 position = position_dodge(width = 0.4),
                 height = 0,
                 linewidth = 1.5) +
 geom_text(
    data        = annot_df_clade,
    aes(x = 0.03, y = 1, label = expr),
    parse       = TRUE,
    inherit.aes = FALSE,
    hjust       = 0,      # 0 = flush against the right edge since x is already max*1.05
    vjust       = 0,      # 0 = flush above the top bar
    size        = 3
  ) +
  apatheme +
  theme(#strip.background = element_rect(fill = NA, colour = "black"),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.line.y = element_blank(),
        strip.text = element_text(size = 10),
        panel.grid.major.x = element_line(colour = "grey90"),
    panel.border = element_rect(colour = "grey90", fill = NA, linewidth = 0.5),
    legend.spacing = unit(0.1, "cm"),  
    legend.key.size = unit(0.3, "cm"),
    legend.margin = ggplot2::margin(0, 0, 0, 0),
    legend.text = element_text(size = 9),
    legend.position = "bottom")

print(imp_plot_pdp_rf_clades)

# prepare rug data
rug_all <- reshape2::melt(clade_results$I$rug) %>%
  #filter(variable %in% unique(sig_lookup$feature)) |>
  #filter(variable %in% clim_vars) |>
  #filter(.feature %in% clim_vars) %>%
  #dplyr::rename(variable = .feature) #%>%
  dplyr::rename(feature = variable) |>
  mutate(var_type = case_when(
    feature %in% c(clim_vars_PC, clim_vars) ~ "Climate",
    feature %in% c(litter_vars, litter_vars_PC) ~ "Litter chemistry",
    TRUE ~ NA
  )) |>
  mutate(feature = gsub("[._]", " ", feature)) |>
  mutate(feature = dplyr::recode(feature,
                       `PC1 clim` = "PC1",
                       `PC2 clim` = "PC2")) |>
  mutate(feature = factor(feature, levels = feature_levels))


    

pdp_plot_ci_clades <- all_pdp_ci %>%
    # 1) classify each feature
    mutate(
      var_type = case_when(
        feature %in% clim_vars_PC  ~ "Climate",
        feature %in% litter_vars   ~ "Litter chemistry",
        TRUE                        ~ NA_character_
      )) |>
    mutate(var_type = factor(var_type ,levels = c("Climate", "Litter chemistry"))) |>
  
      # 2) clean up any underscores
    mutate(
      feature = gsub("[_.]", " ", feature)) |> 
  
      # 3) optionally recode the feature names
      mutate(
        feature = dplyr::recode(feature,
                       `PC1 clim` = "PC1",
                       `PC2 clim` = "PC2")) |>
      mutate(feature = factor(feature, levels = feature_levels)
           ) |> 
      # 4) Optionally remove "UND" (undefined) clade
      dplyr::filter(!Clade %in% c("UND", "V")) |> 
  
    # 4) Generate plot
    ggplot(aes(x = value, y = mean_pred, color = Clade, linetype = significant)) +
    
      # 4) ribbon + line
      #geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = feature), alpha = 0.2, colour = NA) +
      geom_line(size = 1) +
    
      # 5) percent y-axis
      scale_y_continuous(labels = percent_format(scale = 100)) +
    
      # 6) match your custom palette
      scale_color_manual(values = clade_colors, limits = c("I", "II", "III", "IV", "UND")) +
      #scale_fill_manual(values = allvars_pallete) +
      scale_linetype_manual(name = NULL,
                            #name = expression(italic("P")),
                            values = c("TRUE" = "solid", "FALSE" = "longdash"),
                            labels = c(expression(P > 0.05),                      # math‐style labels
                                       expression(P < 0.05))) +
    
      # 7) show raw data distribution on the x-axis
      geom_rug(data = rug_all, aes(x = value), sides = "b", inherit.aes = FALSE) +
    
      # 8) nested facets by type + feature
      ggh4x::facet_nested(~ var_type + feature, scales = "free_x") +
    
      # 9) labels
      labs(
        x     = "Variable value",
        y     = paste("Predicted relative abundance")
      ) +
    
      # 10) drop the legend (we encode feature in the facet strip)
      guides(color = "none",
             fill = "none",
             linetype = guide_legend(order = 1,
                        override.aes = list(size = 1))
                        ) +
    
      # 11) your custom theme + tweaks
      apatheme +
      theme(
        axis.title.y    = element_text(size = 8, margin = ggplot2::margin(r = 10)),
        axis.text.y     = element_text(size = 8),
        strip.background = element_rect(fill = NA, colour = "black"),
        strip.text      = element_text(size = 10),
        panel.grid.major.y = element_line(colour = "grey90"),
        text            = element_text(size = 10),
        # position legend inside top-left of each panel
        legend.position     = c(0.02, 0.98),
        legend.justification = c(0, 1),
        legend.background   = element_rect(fill = alpha("white", 0.7), colour = "black"),
        legend.key.size     = unit(0.3, "cm"),
        legend.text         = element_text(size = 6),
        legend.key.width = unit(1, "cm"),   # give more horizontal room for the dash
        legend.key.height= unit(0.5, "cm"),
        legend.title = element_text(size = 8)
      )

print(pdp_plot_ci_clades)



ggarrange(pdp_plot_ci_clades, imp_plot_pdp_rf_clades,
          labels = "AUTO",
          ncol = 2,
          common.legend = F,
          legend = "bottom",
          widths = c(1.5, 0.5),
          hjust = -0.05, vjust = 1)

width <- 240

ggsave(filename = here::here("plots/plots_pdp/pdp_plot_rf_clades_PCclim_litt_vars.png"),
       width =  width,
       height = width*0.5,
       units = "mm",
       dpi = 300)




```

##### PDP clade level with generalized linear model model

```{r pdp-clades-glm, message=FALSE, warning=FALSE, results='hold', echo=FALSE, fig.width=25, fig.height=20,}
set.seed(711)

# Set up an output directory to save the plots
output_dir <- here::here("plots/plots_glm_pdp")
#dir.create(output_dir, showWarnings = FALSE)
 
#initialize empty data frames to store results
imp_results <- data.frame()
effs_results <- data.frame()

#Set a loop to go over each clade
for (clade in unique(clades_Ab_md$Clade)) {
  #Set a loop to go over each type of litter
  #for (litter in unique(clades_Ab_md$Litter_type)) {
    #subset for each clade and each type of litter
    pdp_temp <- clades_Ab_md %>%
      filter(Clade == clade #& Litter_type == litter
                                ) %>%
      ungroup() %>%
      dplyr::select(-c(Row.names, reserve_code, Litter_type, Clade)) |>
      dplyr::select(all_of(c(litter_vars, clim_vars_PC)), Abundance)
    #create a GLM for each subseted dataset
    glm.mod <- glm(Abundance ~., data = pdp_temp)
    #get the variables for Abundace prediction
    X <- pdp_temp[!(names(pdp_temp) %in% "Abundance")]
    #set model variables for prediction of Abundance
    predictor <- Predictor$new(glm.mod, data = X, y = pdp_temp$Abundance)
    #obtain importance of varibles in the model
    imp <- FeatureImp$new(predictor, loss = "rmse", n.repetitions = 100, compare = "difference")
    
    #store importance results into a data frame and add clade and litter type information
    imp_temp <-imp$results
    imp_temp$Clade <- clade
    #imp_temp$Litter_type <- litter
    
    imp_results <- rbind(imp_results, imp_temp)
    
    #calculate predictions
    effs <- FeatureEffects$new(predictor, grid.size = 50, method = "pdp")
    effs_temp <- do.call(rbind, effs$results)
    effs_temp$Clade <- clade
    #effs_temp$Litter_type <- litter
   
    effs_results <- rbind(effs_results, effs_temp)
  #}
}

imp_plot_pdp_glm <- imp_results %>%
  mutate(var_type = case_when(feature %in% clim_vars_PC ~ "Climate",
                              feature %in% litter_vars ~ "Litter chem.",
                              TRUE ~ "All")) |>
  mutate(feature = gsub("\\.", " ", feature)) %>%
    ggplot(aes(x = importance, y = feature, color = Clade)) +
    scale_color_manual(values = clade_colors, limits = c("I", "II", "III", "IV", "V", "UND")) +
    facet_grid(var_type~., scales = "free_y", space = "free_y") +
    geom_point(position = position_dodge(width = 0.4), size = 3) +
    geom_errorbarh(aes(xmin = importance.05, xmax = importance.95),
                   position = position_dodge(width = 0.4),
                   height = 0,
                   linewidth = 1.5) +
  apatheme +
  theme(text = element_text(size = 20),
    strip.background = element_rect(fill = NA, colour = "black"),
                                                            strip.text = element_text(size = 10),
                                                            panel.grid.major.x = element_line(colour = "grey90"))
print(imp_plot_pdp_glm)

pdp_plot_glm <- effs_results %>%
  mutate(var_type = case_when(.feature %in% clim_vars_PC ~ "Climate",
                              .feature %in% litter_vars ~ "Litter chem.",
                              TRUE ~ NA)) |>
  mutate(.feature = gsub("\\.", " ", .feature)) %>%
  ggplot(aes(x = .borders, y = .value, color = Clade)) +
  scale_color_manual(values = clade_colors, limits = c("I", "II", "III", "IV", "V", "UND")) +
  scale_y_continuous(labels = scales::percent_format(scale = 0.1)) +
  geom_rug(data = rug_all, aes(x = value), sides = "b", inherit.aes = F) +
  #geom_rug(data = rug_G, aes(x = value), sides = "b", inherit.aes = F) +
  #facet_grid(Litter_type~.feature, scales = "free_x") +
  ggh4x::facet_nested(~var_type + .feature, scales = "free_x") +
  labs(x = "Variable value", y = "Predicted relative abundance") +
  geom_line(linewidth = 1) +
  apatheme +
  theme(text = element_text(size = 18),
    strip.background = element_rect(fill = NA, colour = "black"),
        strip.text = element_text(size = 10),
        panel.grid.major.y = element_line(colour = "grey90"))

print(pdp_plot_glm)

ggarrange(pdp_plot_glm, imp_plot_pdp_glm,
          labels = "AUTO",
          ncol = 2,
          common.legend = T,
          legend = "bottom",
          widths = c(1, 0.5))

ggsave(filename = here::here("plots_glm_pdp/pdp_plot_glm_clades.png"),
       width =  65,
       height = 25,
       units = "cm",
       dpi = 320)


##### climate PCs


#initialize empty data frames to store results
imp_results <- data.frame()
effs_results <- data.frame()

#Set a loop to go over each clade
for (clade in unique(clades_Ab_md$Clade)) {
  #Set a loop to go over each type of litter
  #for (litter in unique(clades_Ab_md$Litter_type)) {
    #subset for each clade and each type of litter
    pdp_temp <- clades_Ab_md %>%
      filter(Clade == clade #& Litter_type == litter
                                ) %>%
      ungroup() %>%
      dplyr::select(Abundance, PC1.clim, PC2.clim)
    #create a GLM for each subseted dataset
    glm.mod <- glm(Abundance ~., data = pdp_temp)
    #get the variables for Abundace prediction
    X <- pdp_temp[!(names(pdp_temp) %in% "Abundance")]
    #set model variables for prediction of Abundance
    predictor <- Predictor$new(glm.mod, data = X, y = pdp_temp$Abundance)
    #obtain importance of varibles in the model
    imp <- FeatureImp$new(predictor, loss = "rmse", n.repetitions = 100, compare = "difference")
    
    #store importance results into a data frame and add clade and litter type information
    imp_temp <-imp$results
    imp_temp$Clade <- clade
    #imp_temp$Litter_type <- litter
    
    imp_results <- rbind(imp_results, imp_temp)
    
    #calculate predictions
    effs <- FeatureEffects$new(predictor, grid.size = 50, method = "pdp")
    effs_temp <- do.call(rbind, effs$results)
    effs_temp$Clade <- clade
    #effs_temp$Litter_type <- litter
   
    effs_results <- rbind(effs_results, effs_temp)
    
    rug <- X
    #object_name <- paste("rug", litter, sep = "_")
    #assign(object_name, X)
  #}
}

rug_pca<- reshape2::melt(rug) %>%
  dplyr::rename(.feature = variable) %>%
  #filter(!.feature %in% c(litter_vars, clim_vars)) %>%
  mutate(.feature = gsub("\\.", " ", .feature))


imp_plot_pdp_glm_pca_clim <- imp_results %>%
  #mutate(var_type = case_when(feature %in% clim_vars ~ "Climate",
  #                            feature %in% litter_vars ~ "Litter chem.",
  #                            TRUE ~ "All")) |>
  mutate(feature = gsub("\\.", " ", feature)) %>%
    ggplot(aes(x = importance, y = feature, color = Clade)) +
    scale_color_manual(values = clade_colors, limits = c("I", "II", "III", "IV", "V", "UND")) +
    #facet_grid(var_type~., scales = "free_y", space = "free_y") +
    geom_point(position = position_dodge(width = 0.4), size = 3) +
    geom_errorbarh(aes(xmin = importance.05, xmax = importance.95),
                   position = position_dodge(width = 0.4),
                   height = 0,
                   linewidth = 1.5) +
  apatheme +
  theme(text = element_text(size = 20),
    strip.background = element_rect(fill = NA, colour = "black"),
                                                            strip.text = element_text(size = 10),
                                                            panel.grid.major.x = element_line(colour = "grey90"))
print(imp_plot_pdp_glm_pca_clim)

pdp_plot_glm_pca_clim <- effs_results %>%
  #mutate(var_type = case_when(.feature %in% clim_vars ~ "Climate",
  #                            .feature %in% litter_vars ~ "Litter chem.",
  #                            TRUE ~ "All")) |>
  mutate(.feature = gsub("\\.", " ", .feature)) %>%
  ggplot(aes(x = .borders, y = .value, color = Clade)) +
  scale_color_manual(values = clade_colors, limits = c("I", "II", "III", "IV", "V", "UND")) +
  scale_y_continuous(labels = scales::percent_format(scale = 0.1)) +
  geom_rug(data = rug_pca, aes(x = value), sides = "b", inherit.aes = F) +
  #geom_rug(data = rug_G, aes(x = value), sides = "b", inherit.aes = F) +
  #facet_grid(Litter_type~.feature, scales = "free_x") +
  facet_grid(~.feature, scales = "free_x") +
  labs(x = "Variable value", y = "Predicted relative abundance") +
  geom_line(linewidth = 1) +
  apatheme +
  theme(text = element_text(size = 18),
    strip.background = element_rect(fill = NA, colour = "black"),
        strip.text = element_text(size = 10),
        panel.grid.major.y = element_line(colour = "grey90"))

print(pdp_plot_glm_pca_clim)

ggarrange(pdp_plot_glm_pca_clim, imp_plot_pdp_glm_pca_clim,
          labels = "AUTO",
          ncol = 2,
          common.legend = T,
          legend = "bottom",
          widths = c(1, 0.5))

ggsave(filename = here::here("plots_glm_pdp/pdp_plot_glm_clades_pca_clim.png"),
       width =  65,
       height = 25,
       units = "cm",
       dpi = 320)



```

##### PDP clade level with run_glm_analysis (bootstrapped CIs)

```{r pdp-clades-glm-ci, message=FALSE, warning=FALSE, results='hold', echo=FALSE, fig.width=25, fig.height=20}
set.seed(711)

clade_glm_results <- cached(here::here("cache/clade_glm_results.rds"), {
  res_list <- list()
  for (clade in unique(clades_Ab_md$Clade)) {
    message("Processing clade: ", clade)
    df_cl <- clades_Ab_md %>%
      filter(Clade == clade) %>%
      dplyr::select(all_of(c("Abundance", litter_vars, clim_vars_PC))) %>%
      na.omit()
    res <- run_glm_analysis(
      df           = df_cl,
      response_var = "Abundance",
      predictors   = c(litter_vars, clim_vars_PC),
      cv_folds = 10, nrep_vi = 500, n_boot = 100, grid.size = 20)
    res$summary_table$Clade  <- clade
    res$pdp_ci_df$Clade      <- clade
    res$imp_results_rf$Clade <- clade
    res_list[[clade]] <- res
  }
  res_list
})

# Combine into single tables
all_summaries_glm <- bind_rows(lapply(clade_glm_results, `[[`, "summary_table"))
all_pdp_ci_glm    <- bind_rows(lapply(clade_glm_results, `[[`, "pdp_ci_df"))
all_imp_glm       <- bind_rows(lapply(clade_glm_results, `[[`, "imp_results_rf"))

# Significant variables per clade (from GLM coefficient p-values)
sig_lookup_glm <- map_dfr(
  .x = names(clade_glm_results),
  .f = function(cl) tibble(Clade = cl, feature = clade_glm_results[[cl]]$sig_vars_rfp)
)

write.table(all_summaries_glm, file = here::here("plots/GLM_validation_clades.csv"),
            quote = F, row.names = F, col.names = T, sep = ",")

all_imp_sig_glm <- all_imp_glm %>% semi_join(sig_lookup_glm, by = c("Clade", "feature"))

# Recode features to match feature_levels
all_imp_sig_glm <- all_imp_sig_glm %>%
  mutate(
    feature = gsub("[._]", " ", feature),
    feature = dplyr::recode(feature, `PC1 clim` = "PC1", `PC2 clim` = "PC2"),
    feature = if_else(feature %in% c("crude protein", "crude_protein"), "crude protein", feature)
  )

# Annotation positions
pos_df_clade_glm <- all_imp_sig_glm %>%
  group_by(Clade) %>%
  reframe(x = max(importance, na.rm = TRUE) * 1.05, y = n() + 0.3)

annot_df_clade_glm <- all_summaries_glm %>%
  filter(Metric %in% c("10-fold CV R²", "Overall model p-value (F-test)")) %>%
  dplyr::select(Clade, Metric, Value) %>%
  pivot_wider(names_from = Metric, values_from = Value) %>%
  dplyr::rename(CV_R2 = `10-fold CV R²`, P_val = `Overall model p-value (F-test)`) %>%
  mutate(
    CV_R2 = as.numeric(CV_R2),
    P_num = as.numeric(sub("^<", "", P_val)),
    expr  = paste0(
      "atop(",
        "CV~R^2==", round(CV_R2, 3), ",",
        "italic(P)==", format.pval(P_num, digits = 2, eps = 1e-3),
      ")"
    )
  ) %>%
  left_join(pos_df_clade_glm, by = "Clade") %>%
  dplyr::filter(!Clade %in% c("V", "UND"))

# Importance plot
imp_plot_pdp_glm_clades <- all_imp_sig_glm %>%
  mutate(var_type = case_when(feature %in% clim_vars_PC ~ "Climate",
                              feature %in% litter_vars  ~ "Litter chemistry",
                              TRUE ~ "All"),
         Clade = factor(Clade, levels = c("I", "II", "III", "IV", "V", "UND"))) |>
  dplyr::filter(Clade != "UND") |>
  mutate(feature = factor(feature, levels = feature_levels_imp)) %>%
  ggplot(aes(x = importance, y = feature, color = feature)) +
  scale_color_manual(values = allvars_pallete_imp,
                     labels = c(PC1 = "PC1 clim", PC2 = "PC2 clim"),
                     guide = guide_legend(ncol = 2)) +
  scale_y_discrete(limits = rev(c("PC1", "PC2", "lignin", "cellulose", "hemicellulose", "crude protein", "pH"))) +
  scale_x_continuous(labels = scales::percent_format()) +
  ggh4x::facet_grid2(Clade ~ ., scales = "free_y", space = "free_y",
                     strip = ggh4x::strip_themed(
                       background_y = ggh4x::elem_list_rect(fill = clade_colors)
                     )) +
  geom_point(position = position_dodge(width = 0.4), size = 2.5) +
  geom_errorbarh(aes(xmin = importance.05, xmax = importance.95),
                 position = position_dodge(width = 0.4), height = 0, linewidth = 1.5) +
  geom_text(data = annot_df_clade_glm,
            aes(x = 0.03, y = 1, label = expr),
            parse = TRUE, inherit.aes = FALSE, hjust = 0, vjust = 0, size = 3) +
  apatheme +
  theme(axis.text.y = element_blank(), axis.title.y = element_blank(),
        axis.line.y = element_blank(), strip.text = element_text(size = 10),
        panel.grid.major.x = element_line(colour = "grey90"),
        panel.border = element_rect(colour = "grey90", fill = NA, linewidth = 0.5),
        legend.spacing = unit(0.1, "cm"), legend.key.size = unit(0.3, "cm"),
        legend.margin = ggplot2::margin(0, 0, 0, 0),
        legend.text = element_text(size = 9), legend.position = "bottom")

print(imp_plot_pdp_glm_clades)

# PDP plot
# Reuse rug_all from pdp-clades-rf (same predictor space)
pdp_plot_ci_clades_glm <- all_pdp_ci_glm %>%
  mutate(var_type = case_when(feature %in% clim_vars_PC ~ "Climate",
                              feature %in% litter_vars  ~ "Litter chemistry",
                              TRUE ~ NA_character_)) |>
  mutate(var_type = factor(var_type, levels = c("Climate", "Litter chemistry"))) |>
  mutate(feature = gsub("[_.]", " ", feature)) |>
  mutate(feature = dplyr::recode(feature, `PC1 clim` = "PC1", `PC2 clim` = "PC2")) |>
  mutate(feature = factor(feature, levels = feature_levels)) |>
  dplyr::filter(!Clade %in% c("UND")) |>
  ggplot(aes(x = value, y = mean_pred, color = Clade, linetype = significant)) +
  geom_line(size = 1) +
  scale_y_continuous(labels = percent_format(scale = 100)) +
  scale_color_manual(values = clade_colors, limits = c("I", "II", "III", "IV", "V")) +
  scale_linetype_manual(name = NULL,
                        values = c("TRUE" = "solid", "FALSE" = "longdash"),
                        labels = c(expression(P > 0.05), expression(P < 0.05))) +
  geom_rug(data = rug_all, aes(x = value), sides = "b", inherit.aes = FALSE) +
  ggh4x::facet_nested(~ var_type + feature, scales = "free_x") +
  labs(x = "Variable value", y = "Predicted relative abundance") +
  guides(color = "none", fill = "none",
         linetype = guide_legend(order = 1, override.aes = list(size = 1))) +
  apatheme +
  theme(axis.title.y = element_text(size = 8, margin = ggplot2::margin(r = 10)),
        axis.text.y = element_text(size = 8),
        strip.background = element_rect(fill = NA, colour = "black"),
        strip.text = element_text(size = 10),
        panel.grid.major.y = element_line(colour = "grey90"),
        text = element_text(size = 10),
        legend.position = c(0.02, 0.98), legend.justification = c(0, 1),
        legend.background = element_rect(fill = alpha("white", 0.7), colour = "black"),
        legend.key.size = unit(0.3, "cm"), legend.text = element_text(size = 6),
        legend.key.width = unit(1, "cm"), legend.key.height = unit(0.5, "cm"),
        legend.title = element_text(size = 8))

print(pdp_plot_ci_clades_glm)

ggarrange(pdp_plot_ci_clades_glm, imp_plot_pdp_glm_clades,
          labels = "AUTO", ncol = 2, common.legend = F,
          legend = "bottom", widths = c(1.5, 0.5), hjust = -0.05, vjust = 1)

ggsave(filename = here::here("plots/plots_pdp/pdp_plot_glm_clades_PCclim_litt_vars.png"),
       width = width, height = width * 0.5, units = "mm", dpi = 300)
```

##### PDP subclade level with Random Forest model

```{r pdp-subclades-rf, message=FALSE, warning=FALSE, results='hold', echo=FALSE, fig.width=25, fig.height=20}

set.seed(711)

colnames(subclades_Ab_md) <- gsub(" ", ".", colnames(subclades_Ab_md))


#initialize empty data frames to store results
imp_results <- data.frame()
effs_results <- data.frame()

#Set a loop to go over each clade
for (subclade in unique(subclades_Ab_md$Subclade)) {
#for (subclade in c("IC", "IIG", "IIIA", "IVA", "IVB", "VA")) {
  #Set a loop to go over each type of litter
  #for (litter in unique(subclades_Ab_md$Litter_type)) {
    #subset for each clade and each type of litter
    pdp_temp <- subclades_Ab_md %>%
      filter(Subclade == subclade #& Litter_type == litter
                                ) %>%
      ungroup() %>%
      dplyr::select(-c(Row.names, reserve_code, Litter_type, Subclade, clade)) |>
      dplyr::select(all_of(c(litter_vars, clim_vars_PC)), Abundance)
    #create a random forest for each subseted dataset
    rf <- randomForest(Abundance ~., data = pdp_temp, ntree = 1000)
    #get the variables for Abundace prediction
    X <- pdp_temp[!(names(pdp_temp) %in% "Abundance")]
    #set model variables for prediction of Abundance
    predictor <- Predictor$new(rf, data = X, y = pdp_temp$Abundance)
    #obtain importance of varibles in the model
    imp <- FeatureImp$new(predictor, loss = "rmse", n.repetitions = 100, compare = "difference")
    
    #store importance results into a data frame and add clade and litter type information
    imp_temp <-imp$results
    imp_temp$Subclade <- subclade
    #imp_temp$Litter_type <- litter
    
    imp_results <- rbind(imp_results, imp_temp)
    
    #calculate predictions
    effs <- FeatureEffects$new(predictor, grid.size = 20, method = "pdp")
    effs_temp <- do.call(rbind, effs$results)
    effs_temp$Subclade <- subclade
    #effs_temp$Litter_type <- litter
   
    effs_results <- rbind(effs_results, effs_temp)
    
    rug <- X
    #object_name <- paste("rug", litter, sep = "_")
    #assign(object_name, X)
  #}
}

subclade_results <- cached(here::here("cache/subclade_results.rds"), {
  res_list <- list()
  for (subclade in unique(subclades_Ab_md$Subclade)) {
    message("Processing subclade: ", subclade)
    df_cl <- subclades_Ab_md %>%
      filter(Subclade == subclade) %>%
      dplyr::select(all_of(c("Abundance", litter_vars, clim_vars_PC))) %>%
      na.omit()
    res <- run_rf_analysis(
      df           = df_cl,
      response_var = "Abundance",
      predictors   = c(litter_vars, clim_vars_PC),
      ntree        = 1000, cv_folds = 10, nperm_oob = 500,
      nrep_vi      = 100, n_boot = 100, grid.size = 20)
    str(res$imp_results_rf)
    res$summary_table$Subclade  <- subclade
    res$pdp_ci_df$Subclade      <- subclade
    res$imp_results_rf$Subclade <- subclade
    res$rf_cv$results$Subclade  <- subclade
    res_list[[subclade]] <- res
  }
  res_list
})

# Now: combine into single tables

# (a) Combined summary of overall metrics:
all_summaries_sub <- bind_rows(
  lapply(subclade_results, `[[`, "summary_table")
)

# (b) Combined PDP+CI for all clades:
all_pdp_ci_sub <- bind_rows(
  lapply(subclade_results, `[[`, "pdp_ci_df")
)

# (c) Combined iml importances:
all_imp_sub <- bind_rows(
  lapply(subclade_results, `[[`, "imp_results_rf")
)

# (d) Combined CV results (if you like):
all_cv_sub <- bind_rows(
  lapply(subclade_results, function(x) x$rf_cv$results)
)


# Make a tibble with one row per clade‐feature pair
sig_lookup_sub <- map_dfr(
  .x   = names(subclade_results),
  .f   = function(cl) {
    tibble(
      Subclade   = cl,
      feature = subclade_results[[cl]]$sig_vars_rfp
    )
  }
)

all_summaries_sub <- all_summaries_sub |>
  pivot_wider(id_cols = Subclade, names_from = "Metric", values_from = "Value")

# Average 10-fold CV R² across subclades whose model is significant (permuted p < 0.05)
mean_cv_r2_sig_sub <- all_summaries_sub |>
  mutate(
    p_val = as.numeric(sub("^<", "", `Permuted CV p-value`)),
    cv_r2 = as.numeric(`10-fold CV R²`)
  ) |>
  dplyr::filter(p_val <= 0.05) |>
  dplyr::summarise(n_significant = dplyr::n(),
                   mean_CV_R2    = mean(cv_r2, na.rm = TRUE))
mean_cv_r2_sig_sub


write.table(all_summaries_sub, file = here::here("plots/RF_validation_subclades.csv"), quote = F, row.names = F, col.names = T, sep = ",")

all_imp_sig_sub    <- all_imp_sub    %>% semi_join(sig_lookup_sub, by = c("Subclade", "feature"))
#all_pdp_ci_sig_sub <- all_pdp_ci_sub %>% semi_join(sig_lookup_sub, by = c("Subclade", "feature"))


# 1) Re‐mutate `all_imp_sig` so that its `feature` column is exactly one of those names:
all_imp_sig_sub <- all_imp_sig_sub %>%
  # 1a) Remove dots/underscores (separate mutate so downstream steps see the result)
  mutate(feature = gsub("[._]", " ", feature)) %>%
  # 1b) Recode the climate PCs and normalise crude protein
  mutate(
    feature = dplyr::recode(feature,
      `PC1 clim` = "PC1",
      `PC2 clim` = "PC2"),
    feature = if_else(feature == "crude protein" | feature == "crude_protein",
                      "crude protein", feature)
    )

# 2) compute finite coords *after* recoding + factoring features
pos_df_subclade <- all_imp_sig_sub %>%
  dplyr::filter(Subclade %in% c("IC", "IIG", "IIIA", "IVB")) |>
  group_by(
    Subclade) %>%
  reframe(
    x = max(importance, na.rm = TRUE) * 1.05,
    y = n() + 0.3           # 0.3 units above the topmost point
  )




# 1) extract per‐clade CV R² & p‐value
annot_df_subclade <- all_summaries_sub %>%
  dplyr::select(Subclade, `10-fold CV R²`, `Permuted CV p-value`) %>%
  dplyr::rename(
    CV_R2    = `10-fold CV R²`,
    P_val    = `Permuted CV p-value`
  ) %>%
  mutate(
    CV_R2 = as.numeric(CV_R2),
    P_num = as.numeric(sub("^<", "", P_val)),
    expr  = paste0(
      "atop(",
        "CV~R^2==", round(CV_R2, 3), ",",
        "italic(P)<", format.pval(P_num, digits=2, eps = 1e-3),
      ")"
    )
  )



annot_df_subclade <- annot_df_subclade %>% left_join(pos_df_subclade, by="Subclade") |>
  dplyr::filter(Subclade %in% c("IC", "IIG", "IIIA", "IVB")) |>
  mutate(Subclade = factor(Subclade, levels = c("IIIA", "IC", "IVB", "IIG")))
  



imp_plot_pdp_rf_sub <- all_imp_sig_sub %>%
  dplyr::filter(Subclade %in% c("IC", "IIG", "IIIA", "IVB")) |>
  mutate(var_type = case_when(feature %in% clim_vars_PC ~ "Climate",
                              feature %in% litter_vars ~ "Litter chemistry",
                              TRUE ~ NA),
         Subclade = factor(Subclade, levels = c("IIIA", "IC", "IVB", "IIG"))) |> 
  mutate(feature = factor(feature, levels = feature_levels_imp)) %>%
  ggplot(aes(x = importance, y = feature, color = feature)) +
  #scale_color_manual(values = subclade_colros, limits = c("IC", "IIG", "IIIA", "IVA", "IVB", "VA")) +
  scale_color_manual(values = allvars_pallete_imp,
                     labels = c(PC1 = "PC1 clim", PC2 = "PC2 clim"),
                     guide = guide_legend(ncol = 2)) +
  scale_y_discrete(limits = rev(c("PC1", "PC2", "lignin", "cellulose", "hemicellulose", "crude protein", "pH"))) +
  scale_x_continuous(labels = scales::percent_format()) +
  #facet_grid(var_type~., scales = "free_y", space = "free_y") +
  ggh4x::facet_grid2(Subclade ~ ., scales = "free_y", space = "fixed",
                       strip = ggh4x::strip_themed(
                         background_y = ggh4x::elem_list_rect(
                           fill = alpha(subclade_colros[names(subclade_colros) %in% c("IIIA", "IC", "IVB", "IIG")][c(3,1,4,2)], 0.75) # one color per Sublade
                         )
                       )) +
  geom_point(position = position_dodge(width = 0.6), size = 3) +
  geom_errorbarh(aes(xmin = importance.05, xmax = importance.95),
                 position = position_dodge(width = 0.6),
                 height = 0,
                 linewidth = 1.5) +
    geom_text(
    data        = annot_df_subclade,
    aes(x = 0.0275, y = 1, label = expr),
    parse       = TRUE,
    inherit.aes = FALSE,
    hjust       = 0,      # 0 = flush against the right edge since x is already max*1.05
    vjust       = 0,      # 0 = flush above the top bar
    size        = 3
  ) +
  xlab("importance (RMSE)") +
   apatheme +
  theme(#strip.background = element_rect(fill = NA, colour = "black"),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.line.y = element_blank(),
        strip.text = element_text(size = 10),
        panel.grid.major.x = element_line(colour = "grey90"),
    panel.border = element_rect(colour = "grey90", fill = NA, linewidth = 0.5),
    legend.title = element_blank(),
    legend.spacing = unit(0.1, "cm"),  
    legend.key.size = unit(0.3, "cm"),
    legend.margin = ggplot2::margin(0, 0, 0, 0),
    legend.text = element_text(size = 11),
    legend.position = "bottom")

print(imp_plot_pdp_rf_sub)


# prepare rug data
rug_all <- reshape2::melt(subclade_results$IIIA$rug) %>%
  #dplyr::filter(variable %in% unique(sig_lookup_sub$feature) & variable != "crude_protein") |>
  dplyr::rename(feature = variable) |>
  mutate(var_type = case_when(
    feature %in% c(clim_vars_PC, clim_vars) ~ "Climate",
    feature %in% c(litter_vars, litter_vars_PC) ~ "Litter chemistry",
    TRUE ~ NA
  )) |>
  mutate(feature = gsub("[._]", " ", feature)) |>
  mutate(feature = dplyr::recode(feature,
                       `PC1 clim` = "PC1",
                       `PC2 clim` = "PC2")) |>
  mutate(feature = factor(feature, levels = feature_levels))

#### plot effects

pdp_plot_ci_subclades <- all_pdp_ci_sub %>%
    # 1) classify each feature
    mutate(
      var_type = case_when(
        feature %in% clim_vars_PC  ~ "Climate",
        feature %in% litter_vars   ~ "Litter chemistry",
        TRUE                        ~ NA_character_
      )) |>
    mutate(var_type = factor(var_type ,levels = c("Climate", "Litter chemistry"))) |>
  
      # 2) clean up any underscores
    mutate(
      feature = gsub("[_.]", " ", feature)) |> 
  
      # 3) optionally recode the feature names
      mutate(
        feature = dplyr::recode(feature,
                       `PC1 clim` = "PC1",
                       `PC2 clim` = "PC2")) |>
      mutate(feature = factor(feature, levels = feature_levels)
           ) |> 
      # 4) Optionally remove "UND" (undefined) clade
      dplyr::filter(Subclade %in% c("IC", "IIG", "IIIA", "IVB")) |> 
  
    # 4) Generate plot
    ggplot(aes(x = value, y = mean_pred, color = Subclade, linetype = significant)) +
    
      # 4) ribbon + line
      #geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = feature), alpha = 0.2, colour = NA) +
      geom_line(size = 1) +
    
      # 5) percent y-axis
      scale_y_continuous(labels = percent_format(scale = 100)) +
    
      # 6) match your custom palette
      scale_color_manual(values = subclade_colors_bp, limits = c("IC", "IIG", "IIIA", "IVB")) +
      #scale_fill_manual(values = allvars_pallete) +
      scale_linetype_manual(name = NULL,
                            #name = expression(italic("P")),
                            values = c("TRUE" = "solid", "FALSE" = "longdash"),
                            labels = c(expression(P > 0.05),                      # math‐style labels
                                       expression(P < 0.05))) +
    
      # 7) show raw data distribution on the x-axis
      geom_rug(data = rug_all, aes(x = value), sides = "b", inherit.aes = FALSE) +
    
      # 8) nested facets by type + feature
      ggh4x::facet_nested(~ var_type + feature, scales = "free_x") +
    
      # 9) labels
      labs(
        x     = "Variable value",
        y     = paste("Predicted relative abundance")
      ) +
    
      # 10) drop the legend (we encode feature in the facet strip)
      guides(color = "none",
             fill = "none",
             linetype = "none") +
    
      # 11) your custom theme + tweaks
      apatheme +
      theme(
        axis.title.y    = element_text(size = 12, margin = ggplot2::margin(r = 10)),
        axis.text.y     = element_text(size = 11),
        axis.text.x     = element_text(size = 8),
        strip.background = element_rect(fill = NA, colour = "black"),
        strip.text      = element_text(size = 10),
        panel.grid.major.y = element_line(colour = "grey90"),
        text            = element_text(size = 10),
        # position legend inside top-left of each panel
        legend.position     = c(0.02, 0.98),
        legend.justification = c(0, 1),
        legend.background   = element_rect(fill = alpha("white", 0.7), colour = "black"),
        legend.key.size     = unit(0.3, "cm"),
        legend.text         = element_text(size = 6),
        legend.key.width = unit(1, "cm"),   # give more horizontal room for the dash
        legend.key.height= unit(0.5, "cm"),
        legend.title = element_text(size = 8)
      )

print(pdp_plot_ci_subclades)

## remove text from importance plot

imp_plot_pdp_rf_sub$layers[[3]] <- NULL


ggarrange(pdp_plot_ci_subclades + theme(plot.margin = ggplot2::margin(5, 15, 5, 5)),
          imp_plot_pdp_rf_sub + theme(plot.margin = ggplot2::margin(5, 5, 5, 15)),
          labels = "AUTO",
          ncol = 2,
          common.legend = F,
          widths = c(1.5, 0.5),
          hjust = -0.05, vjust = 1)

wdith <- 250

ggsave(filename = here::here("plots/plots_pdp/pdp_plot_rf_subclades_PCAclim_litt_vars.png"),
       width =  width,
       height = width*0.5,
       units = "mm",
       dpi = 300,
       bg = "white")


#### low abundance subclades


correct_order[-c(1,4,5,12,16,17,23,25,29)]




imp_plot_pdp_rf_sub_lowAb <- all_imp_sig_sub %>%
  dplyr::filter(!Subclade %in% c("IC", "IIG", "IIIA", "IVB")) |>
  mutate(var_type = case_when(feature %in% clim_vars_PC ~ "Climate",
                              feature %in% litter_vars ~ "Litter chemistry",
                              TRUE ~ NA),
         Subclade = factor(Subclade, levels = correct_order[-c(1,4,5,12,16,17,23,25,29)])) |> 
  mutate(feature = factor(feature, levels = feature_levels_imp)) %>%
  ggplot(aes(x = importance, y = feature, color = feature)) +
  #scale_color_manual(values = subclade_colros, limits = c("IC", "IIG", "IIIA", "IVA", "IVB", "VA")) +
  scale_color_manual(values = allvars_pallete_imp,
                     labels = c(PC1 = "PC1 clim", PC2 = "PC2 clim"),
                     guide = guide_legend(ncol = 2)) +
  scale_y_discrete(limits = rev(c("PC1", "PC2", "lignin", "cellulose", "hemicellulose", "crude protein", "pH"))) +
  scale_x_continuous(labels = scales::percent_format()) +
  #facet_grid(var_type~., scales = "free_y", space = "free_y") +
  ggh4x::facet_grid2(Subclade ~ ., scales = "free_y", space = "free_y",
                       strip = ggh4x::strip_themed(
                         background_y = ggh4x::elem_list_rect(
                           fill = subclade_colors_bp[names(subclade_colors_bp) %in% correct_order[-c(1,4,5,10,12,16,17,23,24,25,29)]] # one color per Sublade
                         )
                       )) +
  geom_point(position = position_dodge(width = 0.6), size = 3) +
  geom_errorbarh(aes(xmin = importance.05, xmax = importance.95),
                 position = position_dodge(width = 0.6),
                 height = 0,
                 linewidth = 1.5) +
  xlab("importance (RMSE)") +
   apatheme +
  theme(#strip.background = element_rect(fill = NA, colour = "black"),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.line.y = element_blank(),
        strip.text = element_text(size = 10),
        panel.grid.major.x = element_line(colour = "grey90"),
    panel.border = element_rect(colour = "grey90", fill = NA, linewidth = 0.5),
    legend.spacing = unit(0.1, "cm"),  
    legend.key.size = unit(0.3, "cm"),
    legend.margin = ggplot2::margin(0, 0, 0, 0),
    legend.text = element_text(size = 9))
print(imp_plot_pdp_rf_sub_lowAb)


#### plot effects


pdp_plot_ci_subclades_lowAb <- all_pdp_ci_sub %>%
  mutate(Clade = case_when(Subclade %in% clade_I_subclades ~ "I",
                           Subclade %in% clade_II_subclades ~ "II",
                           Subclade %in% clade_III_subclades ~ "III",
                           Subclade %in% clade_IV_subclades ~ "IV",
                           Subclade %in% clade_V_subclades ~ "V"
                           )) %>%
    # 1) classify each feature
    mutate(
      var_type = case_when(
        feature %in% clim_vars_PC  ~ "Climate",
        feature %in% litter_vars   ~ "Litter chemistry",
        TRUE                        ~ NA_character_
      )) |>
    mutate(var_type = factor(var_type ,levels = c("Climate", "Litter chemistry"))) |>
  
      # 2) clean up any underscores
    mutate(
      feature = gsub("[_.]", " ", feature)) |> 
  
      # 3) optionally recode the feature names
      mutate(
        feature = dplyr::recode(feature,
                       `PC1 clim` = "PC1",
                       `PC2 clim` = "PC2")) |>
      mutate(feature = factor(feature, levels = feature_levels)
           ) |> 
      # 4) Optionally remove "UND" (undefined) clade
      dplyr::filter(!Subclade %in% c("IC", "IIG", "IIIA", "IVB")) |> 
  
    # 4) Generate plot
    ggplot(aes(x = value, y = mean_pred, color = Subclade, linetype = significant)) +
    
      # 4) ribbon + line
      #geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = feature), alpha = 0.2, colour = NA) +
      geom_line(size = 1) +
    
      # 5) percent y-axis
      scale_y_continuous(labels = percent_format(scale = 100)) +
    
      # 6) match your custom palette
      scale_color_manual(values = subclade_colors_bp) +
      #scale_fill_manual(values = allvars_pallete) +
      scale_linetype_manual(name = NULL,
                            #name = expression(italic("P")),
                            values = c("TRUE" = "solid", "FALSE" = "longdash"),
                            labels = c(expression(P > 0.05),                      # math‐style labels
                                       expression(P < 0.05))) +
    
      # 7) show raw data distribution on the x-axis
      geom_rug(data = rug_all, aes(x = value), sides = "b", inherit.aes = FALSE) +
    
      # 8) nested facets by type + feature
      ggh4x::facet_nested(Clade~var_type + feature, scales = "free_x") +
    
      # 9) labels
      labs(
        x     = "Variable value",
        y     = paste("Predicted relative abundance")
      ) +
    
      # 10) drop the legend (we encode feature in the facet strip)
      guides(color = "none",
             fill = "none",
             linetype = guide_legend(order = 1,
                        override.aes = list(size = 1))
                        ) +
    
      # 11) your custom theme + tweaks
      apatheme +
      theme(
        axis.title.y    = element_text(size = 8, margin = ggplot2::margin(r = 10)),
        axis.text.y     = element_text(size = 8),
        strip.background = element_rect(fill = NA, colour = "black"),
        strip.text      = element_text(size = 10),
        panel.grid.major.y = element_line(colour = "grey90"),
        text            = element_text(size = 10),
        # position legend inside top-left of each panel
        legend.position     = c(0.02, 0.98),
        legend.justification = c(0, 1),
        legend.background   = element_rect(fill = alpha("white", 0.7), colour = "black"),
        legend.key.size     = unit(0.3, "cm"),
        legend.text         = element_text(size = 6),
        legend.key.width = unit(1, "cm"),   # give more horizontal room for the dash
        legend.key.height= unit(0.5, "cm"),
        legend.title = element_text(size = 8)
      )

print(pdp_plot_ci_subclades_lowAb)


ggarrange(pdp_plot_ci_subclades_lowAb, imp_plot_pdp_rf_sub_lowAb,
          labels = "AUTO",
          ncol = 2,
          common.legend = F,
          legend = "bottom",
          widths = c(1.5, 0.5),
          hjust = -0.05, vjust = 1)

wdith <- 260

ggsave(filename = here::here("plots/plots_pdp/pdp_plot_rf_subclades_PCAclim_litt_vars_lowAb.png"),
       width =  width,
       height = width*0.75,
       units = "mm",
       dpi = 300)


```

##### PDP subclade level with generalized linear model model

```{r pdp-subclades-glm, message=FALSE, warning=FALSE, results='hold', echo=FALSE, fig.width=25, fig.height=20,}
set.seed(711)
 
#initialize empty data frames to store results
imp_results <- data.frame()
effs_results <- data.frame()

#Set a loop to go over each clade
#for (subclade in unique(subclades_Ab_md$Subclade)) {
for (subclade in c("IC", "IIG", "IIIA", "IVB", "VA")) {
  #Set a loop to go over each type of litter
  #for (litter in unique(subclades_Ab_md$Litter_type)) {
    #subset for each clade and each type of litter
    pdp_temp <- subclades_Ab_md %>%
      filter(Subclade == subclade #& Litter_type == litter
                                ) %>%
      ungroup() %>%
      dplyr::select(!c(Row.names, reserve_code, Litter_type, Subclade, clade))
    #create a GLM for each subseted dataset
    glm.mod <- glm(Abundance ~., data = pdp_temp)
    #get the variables for Abundace prediction
    X <- pdp_temp[!(names(pdp_temp) %in% "Abundance")]
    #set model variables for prediction of Abundance
    predictor <- Predictor$new(glm.mod, data = X, y = pdp_temp$Abundance)
    #obtain importance of varibles in the model
    imp <- FeatureImp$new(predictor, loss = "rmse", n.repetitions = 100, compare = "difference")
    
    #store importance results into a data frame and add clade and litter type information
    imp_temp <-imp$results
    imp_temp$Subclade <- subclade
    #imp_temp$Litter_type <- litter
    
    imp_results <- rbind(imp_results, imp_temp)
    
    #calculate predictions
    effs <- FeatureEffects$new(predictor, grid.size = 1000, method = "pdp")
    effs_temp <- do.call(rbind, effs$results)
    effs_temp$Subclade <- subclade
    #effs_temp$Litter_type <- litter
   
    effs_results <- rbind(effs_results, effs_temp)
  #}
}

imp_plot_pdp_glm_sub <- imp_results %>%
  mutate(var_type = case_when(feature %in% clim_vars ~ "Climate",
                              feature %in% litter_vars ~ "Litter chem.",
                              TRUE ~ "All")) |>
  mutate(feature = gsub("\\.", " ", feature)) %>%
    ggplot(aes(x = importance, y = feature, color = Subclade)) +
    scale_color_manual(values = subclade_colros, limits = c("IC", "IIG", "IIIA", "IVB", "VA")) +
    #facet_grid(Litter_type~.) +
    facet_grid(var_type~., scales = "free_y", space = "free_y") +
    geom_point(position = position_dodge(width = 0.4), size = 3) +
    geom_errorbarh(aes(xmin = importance.05, xmax = importance.95),
                   position = position_dodge(width = 0.4),
                   height = 0,
                   linewidth = 1.5) + apatheme + theme(strip.background = element_rect(fill = NA, colour = "black"),
                                                            strip.text = element_text(size = 10),
                                                            panel.grid.major.x = element_line(colour = "grey90"))
print(imp_plot_pdp_glm_sub)

pdp_plot_glm_sub <- effs_results %>%
  mutate(var_type = case_when(.feature %in% clim_vars ~ "Climate",
                              .feature %in% litter_vars ~ "Litter chem.",
                              TRUE ~ "All")) |>
  mutate(.feature = gsub("\\.", " ", .feature)) %>%
  ggplot(aes(x = .borders, y = .value, color = Subclade)) +
  scale_color_manual(values = subclade_colros, limits = c("IC", "IIG", "IIIA", "IVB", "VA")) +
  scale_y_continuous(limits = c(0, 10),
                     labels = scales::percent_format(scale = 0.1)) +
  geom_rug(data = rug_all, aes(x = value), sides = "b", inherit.aes = F) +
  #geom_rug(data = rug_G, aes(x = value), sides = "b", inherit.aes = F) +
  #facet_grid(Litter_type~.feature, scales = "free_x") +
  ggh4x::facet_nested(~var_type + .feature, scales = "free_x") +
  labs(x = "Variable value", y = "Predicted relative abundance") +
  geom_line(linewidth = 1) +
  apatheme +
  theme(strip.background = element_rect(fill = NA, colour = "black"),
        strip.text = element_text(size = 10),
        panel.grid.major.y = element_line(colour = "grey90"))

print(pdp_plot_glm_sub)

ggarrange(pdp_plot_glm_sub, imp_plot_pdp_glm_sub,
          labels = "AUTO",
          ncol = 2,
          common.legend = T,
          legend = "bottom",
          widths = c(1, 0.5))

ggsave(filename = here::here("plots_glm_pdp/pdp_plot_glm_subclades.png"),
       width =  65,
       height = 25,
       units = "cm",
       dpi = 320)

```
##### PDP subclade level with run_glm_analysis (bootstrapped CIs)

```{r pdp-subclades-glm-ci, message=FALSE, warning=FALSE, results='hold', echo=FALSE, fig.width=25, fig.height=20}
set.seed(711)

subclade_glm_results <- cached(here::here("cache/subclade_glm_results.rds"), {
  res_list <- list()
  for (subclade in unique(subclades_Ab_md$Subclade)) {
    message("Processing subclade: ", subclade)
    df_cl <- subclades_Ab_md %>%
      filter(Subclade == subclade) %>%
      dplyr::select(all_of(c("Abundance", litter_vars, clim_vars_PC))) %>%
      na.omit()
    res <- run_glm_analysis(
      df           = df_cl,
      response_var = "Abundance",
      predictors   = c(litter_vars, clim_vars_PC),
      cv_folds = 10, nrep_vi = 100, n_boot = 100, grid.size = 20)
    res$summary_table$Subclade  <- subclade
    res$pdp_ci_df$Subclade      <- subclade
    res$imp_results_rf$Subclade <- subclade
    res_list[[subclade]] <- res
  }
  res_list
})

# Combine into single tables
all_summaries_glm_sub <- bind_rows(lapply(subclade_glm_results, `[[`, "summary_table"))
all_pdp_ci_glm_sub    <- bind_rows(lapply(subclade_glm_results, `[[`, "pdp_ci_df"))
all_imp_glm_sub       <- bind_rows(lapply(subclade_glm_results, `[[`, "imp_results_rf"))

# Significant variables per subclade
sig_lookup_glm_sub <- map_dfr(
  .x = names(subclade_glm_results),
  .f = function(cl) tibble(Subclade = cl, feature = subclade_glm_results[[cl]]$sig_vars_rfp)
)

write.table(all_summaries_glm_sub, file = here::here("plots/GLM_validation_subclades.csv"),
            quote = F, row.names = F, col.names = T, sep = ",")

all_imp_sig_glm_sub <- all_imp_glm_sub %>% semi_join(sig_lookup_glm_sub, by = c("Subclade", "feature"))

# Recode features
all_imp_sig_glm_sub <- all_imp_sig_glm_sub %>%
  mutate(
    feature = gsub("[._]", " ", feature),
    feature = dplyr::recode(feature, `PC1 clim` = "PC1", `PC2 clim` = "PC2"),
    feature = if_else(feature %in% c("crude protein", "crude_protein"), "crude protein", feature)
  )

# Annotation positions (high-abundance subclades only)
pos_df_subclade_glm <- all_imp_sig_glm_sub %>%
  dplyr::filter(Subclade %in% c("IC", "IIG", "IIIA", "IVB")) |>
  group_by(Subclade) %>%
  reframe(x = max(importance, na.rm = TRUE) * 1.05, y = n() + 0.3)

annot_df_subclade_glm <- all_summaries_glm_sub %>%
  filter(Metric %in% c("10-fold CV R²", "Overall model p-value (F-test)")) %>%
  dplyr::select(Subclade, Metric, Value) %>%
  pivot_wider(names_from = Metric, values_from = Value) %>%
  dplyr::rename(CV_R2 = `10-fold CV R²`, P_val = `Overall model p-value (F-test)`) %>%
  mutate(
    CV_R2 = as.numeric(CV_R2),
    P_num = as.numeric(sub("^<", "", P_val)),
    expr  = paste0(
      "atop(",
        "CV~R^2==", round(CV_R2, 3), ",",
        "italic(P)==", format.pval(P_num, digits = 2, eps = 1e-3),
      ")"
    )
  ) %>%
  left_join(pos_df_subclade_glm, by = "Subclade") %>%
  dplyr::filter(Subclade %in% c("IC", "IIG", "IIIA", "IVB")) %>%
  mutate(Subclade = factor(Subclade, levels = c("IIIA", "IC", "IVB", "IIG")))

# Rug data (reuse same predictor space as RF)
rug_all_glm_sub <- reshape2::melt(subclade_glm_results$IIIA$rug) %>%
  dplyr::rename(feature = variable) |>
  mutate(var_type = case_when(
    feature %in% c(clim_vars_PC, clim_vars) ~ "Climate",
    feature %in% c(litter_vars, litter_vars_PC) ~ "Litter chemistry",
    TRUE ~ NA
  )) |>
  mutate(feature = gsub("[._]", " ", feature)) |>
  mutate(feature = dplyr::recode(feature, `PC1 clim` = "PC1", `PC2 clim` = "PC2")) |>
  mutate(feature = factor(feature, levels = feature_levels))

# --- High-abundance subclades ---

imp_plot_pdp_glm_sub <- all_imp_sig_glm_sub %>%
  dplyr::filter(Subclade %in% c("IIIA", "IC", "IVB", "IIG")) |>
  # Add placeholder rows for subclades with no significant variables so their
  # facet panels still render (empty panels with space = "free_y" have 0 height)
  bind_rows(
    tibble(Subclade = c("IC", "IIG"), feature = "PC1",
           importance = NA_real_, importance.05 = NA_real_, importance.95 = NA_real_)
  ) |>
  mutate(var_type = case_when(feature %in% clim_vars_PC ~ "Climate",
                              feature %in% litter_vars  ~ "Litter chemistry",
                              TRUE ~ NA),
         Subclade = factor(Subclade, levels = c("IIIA", "IC", "IVB", "IIG"))) |>
  mutate(feature = dplyr::recode(feature, `PC1 clim` = "PC1", `PC2 clim` = "PC2")) |>
  mutate(feature = factor(feature, levels = feature_levels_imp)) %>%
  ggplot(aes(x = importance, y = feature, color = feature)) +
  scale_color_manual(values = allvars_pallete_imp,
                     labels = c(PC1 = "PC1 clim", PC2 = "PC2 clim"),
                     guide = guide_legend(ncol = 2)) +
  scale_y_discrete(limits = rev(c("PC1", "PC2", "lignin", "cellulose", "hemicellulose", "crude protein", "pH"))) +
  scale_x_continuous(labels = scales::percent_format()) +
  ggh4x::facet_grid2(Subclade ~ ., scales = "free_y", space = "free_y",
                     strip = ggh4x::strip_themed(
                       background_y = ggh4x::elem_list_rect(
                         fill = subclade_colros[c("IIIA", "IC", "IVB", "IIG")]
                       )
                     )) +
  geom_point(position = position_dodge(width = 0.6), size = 3) +
  geom_errorbarh(aes(xmin = importance.05, xmax = importance.95),
                 position = position_dodge(width = 0.6), height = 0, linewidth = 1.5) +
  geom_text(data = annot_df_subclade_glm |> dplyr::filter(Subclade %in% c("IIIA", "IC", "IVB", "IIG")),
            aes(x = 0.0275, y = 1, label = expr),
            parse = TRUE, inherit.aes = FALSE, hjust = 0, vjust = 0, size = 3) +
  xlab("importance (RMSE)") +
  apatheme +
  theme(axis.text.y = element_blank(), axis.title.y = element_blank(),
        axis.line.y = element_blank(), strip.text = element_text(size = 10),
        panel.grid.major.x = element_line(colour = "grey90"),
        panel.border = element_rect(colour = "grey90", fill = NA, linewidth = 0.5),
        legend.spacing = unit(0.1, "cm"), legend.key.size = unit(0.3, "cm"),
        legend.margin = ggplot2::margin(0, 0, 0, 0),
        legend.text = element_text(size = 9), legend.position = "bottom",
        plot.margin = ggplot2::margin(t = 5, r = 5, b = 5, l = 15))

print(imp_plot_pdp_glm_sub)

pdp_plot_ci_subclades_glm <- all_pdp_ci_glm_sub %>%
  mutate(var_type = case_when(feature %in% clim_vars_PC ~ "Climate",
                              feature %in% litter_vars  ~ "Litter chemistry",
                              TRUE ~ NA_character_)) |>
  mutate(var_type = factor(var_type, levels = c("Climate", "Litter chemistry"))) |>
  mutate(feature = gsub("[_.]", " ", feature)) |>
  mutate(feature = dplyr::recode(feature, `PC1 clim` = "PC1", `PC2 clim` = "PC2")) |>
  mutate(feature = factor(feature, levels = feature_levels)) |>
  dplyr::filter(Subclade %in% c("IC", "IIG", "IIIA", "IVB")) |>
  ggplot(aes(x = value, y = mean_pred, color = Subclade, linetype = significant)) +
  geom_line(size = 1) +
  scale_y_continuous(labels = percent_format(scale = 100)) +
  scale_color_manual(values = subclade_colors_bp, limits = c("IC", "IIG", "IIIA", "IVB")) +
  scale_linetype_manual(name = NULL,
                        values = c("TRUE" = "solid", "FALSE" = "longdash"),
                        labels = c(expression(P > 0.05), expression(P < 0.05))) +
  geom_rug(data = rug_all_glm_sub, aes(x = value), sides = "b", inherit.aes = FALSE) +
  ggh4x::facet_nested(~ var_type + feature, scales = "free_x") +
  labs(x = "Variable value", y = "Predicted relative abundance") +
  guides(color = "none", fill = "none",
         linetype = guide_legend(order = 1, override.aes = list(size = 1))) +
  apatheme +
  theme(axis.title.y = element_text(size = 8, margin = ggplot2::margin(r = 10)),
        axis.text.y = element_text(size = 8), axis.text.x = element_text(size = 8),
        strip.background = element_rect(fill = NA, colour = "black"),
        strip.text = element_text(size = 10),
        panel.grid.major.y = element_line(colour = "grey90"),
        text = element_text(size = 10),
        legend.position = c(0.02, 0.98), legend.justification = c(0, 1),
        legend.background = element_rect(fill = alpha("white", 0.7), colour = "black"),
        legend.key.size = unit(0.3, "cm"), legend.text = element_text(size = 6),
        legend.key.width = unit(1, "cm"), legend.key.height = unit(0.5, "cm"),
        legend.title = element_text(size = 8))

print(pdp_plot_ci_subclades_glm)

imp_plot_pdp_glm_sub$layers[[3]] <- NULL  # remove annotation layer for combined plot

ggarrange(pdp_plot_ci_subclades_glm, imp_plot_pdp_glm_sub,
          labels = "AUTO", ncol = 2, common.legend = F,
          legend = "bottom", widths = c(1.5, 0.5), hjust = -0.05, vjust = 1)

ggsave(filename = here::here("plots/plots_pdp/pdp_plot_glm_subclades_PCAclim_litt_vars.png"),
       width = width, height = width * 0.5, units = "mm", dpi = 300, bg = "white")

# --- Low-abundance subclades ---

# Ordered low-abundance subclades present in the data — used for both factor levels and strip colors
subs_lowAb <- intersect(
  correct_order[-c(1, 4, 5, 12, 16, 17, 23, 25, 29)],
  unique(all_imp_sig_glm_sub$Subclade[!all_imp_sig_glm_sub$Subclade %in% c("IC", "IIG", "IIIA", "IVB")])
)

imp_plot_pdp_glm_sub_lowAb <- all_imp_sig_glm_sub %>%
  dplyr::filter(!Subclade %in% c("IC", "IIG", "IIIA", "IVB")) |>
  mutate(var_type = case_when(feature %in% clim_vars_PC ~ "Climate",
                              feature %in% litter_vars  ~ "Litter chemistry",
                              TRUE ~ NA),
         Subclade = factor(Subclade, levels = subs_lowAb)) |>
  mutate(feature = factor(feature, levels = feature_levels_imp)) %>%
  ggplot(aes(x = importance, y = feature, color = feature)) +
  scale_color_manual(values = allvars_pallete_imp,
                     labels = c(PC1 = "PC1 clim", PC2 = "PC2 clim"),
                     guide = guide_legend(ncol = 2)) +
  scale_y_discrete(limits = rev(c("PC1", "PC2", "lignin", "cellulose", "hemicellulose", "crude protein", "pH"))) +
  scale_x_continuous(labels = scales::percent_format()) +
  ggh4x::facet_grid2(Subclade ~ ., scales = "free_y", space = "free_y",
                     strip = ggh4x::strip_themed(
                       background_y = ggh4x::elem_list_rect(
                         fill = subclade_colors_bp[subs_lowAb]
                       )
                     )) +
  geom_point(position = position_dodge(width = 0.6), size = 3) +
  geom_errorbarh(aes(xmin = importance.05, xmax = importance.95),
                 position = position_dodge(width = 0.6), height = 0, linewidth = 1.5) +
  xlab("importance (RMSE)") +
  apatheme +
  theme(axis.text.y = element_blank(), axis.title.y = element_blank(),
        axis.line.y = element_blank(), strip.text = element_text(size = 10),
        panel.grid.major.x = element_line(colour = "grey90"),
        panel.border = element_rect(colour = "grey90", fill = NA, linewidth = 0.5),
        legend.spacing = unit(0.1, "cm"), legend.key.size = unit(0.3, "cm"),
        legend.margin = ggplot2::margin(0, 0, 0, 0),
        legend.text = element_text(size = 9))

print(imp_plot_pdp_glm_sub_lowAb)

 pdp_plot_ci_subclades_glm_lowAb <- all_pdp_ci_glm_sub %>%
  mutate(Clade = case_when(Subclade %in% clade_I_subclades   ~ "I",
                           Subclade %in% clade_II_subclades  ~ "II",
                           Subclade %in% clade_III_subclades ~ "III",
                           Subclade %in% clade_IV_subclades  ~ "IV",
                           Subclade %in% clade_V_subclades   ~ "V")) %>%
  mutate(var_type = case_when(feature %in% clim_vars_PC ~ "Climate",
                              feature %in% litter_vars  ~ "Litter chemistry",
                              TRUE ~ NA_character_)) |>
  mutate(var_type = factor(var_type, levels = c("Climate", "Litter chemistry"))) |>
  mutate(feature = gsub("[_.]", " ", feature)) |>
  mutate(feature = dplyr::recode(feature, `PC1 clim` = "PC1", `PC2 clim` = "PC2")) |>
  mutate(feature = factor(feature, levels = feature_levels)) |>
  dplyr::filter(!Subclade %in% c("IC", "IIG", "IIIA", "IVB")) |>
  ggplot(aes(x = value, y = mean_pred, color = Subclade, linetype = significant)) +
  geom_line(size = 1) +
  scale_y_continuous(labels = percent_format(scale = 100)) +
  scale_color_manual(values = subclade_colors_bp) +
  scale_linetype_manual(name = NULL,
                        values = c("TRUE" = "solid", "FALSE" = "longdash"),
                        labels = c(expression(P > 0.05), expression(P < 0.05))) +
  geom_rug(data = rug_all_glm_sub, aes(x = value), sides = "b", inherit.aes = FALSE) +
  ggh4x::facet_nested(Clade ~ var_type + feature, scales = "free_x") +
  labs(x = "Variable value", y = "Predicted relative abundance") +
  guides(color = "none", fill = "none",
         linetype = guide_legend(order = 1, override.aes = list(size = 1))) +
  apatheme +
  theme(axis.title.y = element_text(size = 8, margin = ggplot2::margin(r = 10)),
        axis.text.y = element_text(size = 8),
        strip.background = element_rect(fill = NA, colour = "black"),
        strip.text = element_text(size = 10),
        panel.grid.major.y = element_line(colour = "grey90"),
        text = element_text(size = 10),
        legend.position = c(0.02, 0.98), legend.justification = c(0, 1),
        legend.background = element_rect(fill = alpha("white", 0.7), colour = "black"),
        legend.key.size = unit(0.3, "cm"), legend.text = element_text(size = 6),
        legend.key.width = unit(1, "cm"), legend.key.height = unit(0.5, "cm"),
        legend.title = element_text(size = 8))

print(pdp_plot_ci_subclades_glm_lowAb)

ggarrange(pdp_plot_ci_subclades_glm_lowAb, imp_plot_pdp_glm_sub_lowAb,
          labels = "AUTO", ncol = 2, common.legend = F,
          legend = "bottom", widths = c(1.5, 0.5), hjust = -0.05, vjust = 1)

ggsave(filename = here::here("plots/plots_pdp/pdp_plot_glm_subclades_PCAclim_litt_vars_lowAb.png"),
       width = width, height = width * 0.75, units = "mm", dpi = 300)
```

##### 2D heatmap PDPs — top 4 subclades (RF)

```{r pdp-subclades-rf-heatmap, message=FALSE, warning=FALSE, results='hold', echo=FALSE, fig.width=20, fig.height=12}
top4_subclades <- c("IIIA", "IC", "IVB", "IIG")

subclades_2d_pdp_rf <- cached(here::here("cache/subclades_2d_pdp_rf.rds"), {
  map_dfr(top4_subclades, function(sub) {
    message("Subclade: ", sub)
    mod  <- subclade_results[[sub]]$rf_full
    X    <- subclade_results[[sub]]$rug
    y    <- subclade_results[[sub]]$rf_full$y
    pred <- Predictor$new(mod, data = X, y = y)

    map_dfr(clim_vars_PC, function(cv) {
      map_dfr(litter_vars, function(lv) {
        message("  ", cv, " x ", lv)
        eff <- FeatureEffect$new(pred, feature = c(cv, lv), method = "pdp", grid.size = 20)
        eff$results |>
          dplyr::rename(clim_value = all_of(cv), litter_value = all_of(lv)) |>
          dplyr::mutate(clim_var = cv, litter_var = lv, Subclade = sub)
      })
    })
  })
})

heatmap_2d_pdp_rf <- function(sub) {
  subclades_2d_pdp_rf |>
    dplyr::filter(Subclade == sub) |>
    mutate(
      clim_var   = dplyr::recode(gsub("[._]", " ", clim_var),
                                  `PC1 clim` = "PC1", `PC2 clim` = "PC2"),
      litter_var = gsub("_", " ", litter_var),
      litter_var = factor(litter_var, levels = gsub("_", " ", litter_vars)),
      clim_var   = factor(clim_var,   levels = c("PC1", "PC2"))
    ) |>
    ggplot(aes(x = litter_value, y = clim_value, fill = .value)) +
    geom_raster(interpolate = TRUE) +
    scale_fill_viridis_c(name = "Predicted\nabundance",
                         labels = scales::percent_format(scale = 100)) +
    facet_grid(clim_var ~ litter_var, scales = "free") +
    labs(x = "Litter chemistry variable value",
         y = "Climate variable value",
         title = sub) +
    apatheme +
    theme(plot.title        = element_text(hjust = 0.5, face = "bold", size = 11,
                                           colour = subclade_colors_bp[sub]),
          strip.background  = element_rect(fill = NA, colour = "black"),
          strip.text        = element_text(size = 8),
          axis.text         = element_text(size = 7),
          axis.title        = element_text(size = 9),
          panel.spacing     = unit(0.3, "cm"),
          legend.key.height = unit(1, "cm"))
}

pdp_heatmaps_rf <- lapply(top4_subclades, heatmap_2d_pdp_rf)

ggarrange(plotlist = pdp_heatmaps_rf, ncol = 2, nrow = 2,
          common.legend = TRUE, legend = "right")

ggsave(here::here("plots/plots_pdp/pdp_heatmap_rf_top4_subclades.png"),
       width = 300, height = 200, units = "mm", dpi = 300, bg = "white")
```

##### 2D heatmap PDPs — significant predictors only (RF)

```{r pdp-subclades-rf-heatmap-sig, message=FALSE, warning=FALSE, results='hold', echo=FALSE, fig.width=16, fig.height=12}

# Build per-subclade list of significant (clim, litter) pairs
sig_pairs <- lapply(setNames(top4_subclades, top4_subclades), function(sub) {
  sig       <- subclade_results[[sub]]$sig_vars_rfp
  sig_clim  <- intersect(sig, clim_vars_PC)
  sig_litt  <- intersect(sig, litter_vars)
  expand.grid(clim_var = sig_clim, litter_var = sig_litt,
              stringsAsFactors = FALSE)
})

# Drop subclades with no valid pairs
sig_pairs <- Filter(function(x) nrow(x) > 0, sig_pairs)

message("Subclades with significant clim × litter pairs: ",
        paste(names(sig_pairs), collapse = ", "))

subclades_2d_pdp_rf_sig <- cached(here::here("cache/subclades_2d_pdp_rf_sig.rds"), {
  map_dfr(names(sig_pairs), function(sub) {
    message("Subclade: ", sub)
    mod  <- subclade_results[[sub]]$rf_full
    X    <- subclade_results[[sub]]$rug
    y    <- subclade_results[[sub]]$rf_full$y
    pred <- Predictor$new(mod, data = X, y = y)

    pmap_dfr(sig_pairs[[sub]], function(clim_var, litter_var) {
      message("  ", clim_var, " x ", litter_var)
      eff <- FeatureEffect$new(pred, feature = c(clim_var, litter_var),
                               method = "pdp", grid.size = 20)
      eff$results |>
        dplyr::rename(clim_value = all_of(clim_var), litter_value = all_of(litter_var)) |>
        dplyr::mutate(clim_var = clim_var, litter_var = litter_var, Subclade = sub)
    })
  })
})

heatmap_2d_pdp_rf_sig <- function(sub) {
  subclades_2d_pdp_rf_sig |>
    dplyr::filter(Subclade == sub, clim_var == "PC2.clim") |>
    mutate(
      litter_var = gsub("_", " ", litter_var),
      litter_var = factor(litter_var, levels = gsub("_", " ", litter_vars))
    ) |>
    ggplot(aes(x = litter_value, y = clim_value, fill = .value)) +
    geom_raster(interpolate = TRUE) +
    scale_fill_viridis_c(name = "Predicted\nabundance",
                         labels = scales::percent_format(scale = 100)) +
    facet_grid(. ~ litter_var, scales = "free_x") +
    labs(x = "Litter chemistry variable value",
         y = "PC2 climate value",
         title = sub) +
    apatheme +
    theme(plot.title        = element_text(hjust = 0.5, face = "bold", size = 11,
                                           colour = subclade_colors_bp[sub]),
          strip.background  = element_rect(fill = NA, colour = "black"),
          strip.text        = element_text(size = 8),
          axis.text         = element_text(size = 7),
          axis.title        = element_text(size = 9),
          panel.spacing     = unit(0.3, "cm"),
          legend.key.height = unit(1, "cm"))
}

pdp_heatmaps_rf_sig <- lapply(names(sig_pairs), heatmap_2d_pdp_rf_sig)

ggarrange(plotlist = pdp_heatmaps_rf_sig,
          ncol = length(pdp_heatmaps_rf_sig), nrow = 1,
          common.legend = TRUE, legend = "right")

ggsave(here::here("plots/plots_pdp/pdp_heatmap_rf_top4_subclades_sig_vars.png"),
       width = 100 * length(pdp_heatmaps_rf_sig), height = 160,
       units = "mm", dpi = 300, bg = "white")
```

#### RF + PDP with MAT & MAP as climate variables

```{r mat-map-setup, include=FALSE}
clim_vars_MATMAP      <- c("MAT", "MAP")
feature_levels_MATMAP <- c("MAT", "MAP", "lignin", "cellulose", "hemicellulose", "crude protein", "pH")
width_mm              <- 240
```

##### Clade level — RF (MAT & MAP)

```{r pdp-clades-rf-mat-map, message=FALSE, warning=FALSE, results='hold', echo=FALSE, fig.width=25, fig.height=20}
set.seed(711)

clade_results_MATMAP <- cached(here::here("cache/clade_results_MATMAP.rds"), {
  res_list <- list()
  for (clade in unique(clades_Ab_md$Clade)) {
    message("Processing clade: ", clade)
    df_cl <- clades_Ab_md %>%
      filter(Clade == clade) %>%
      dplyr::select(all_of(c("Abundance", litter_vars, clim_vars_MATMAP))) %>%
      na.omit()
    res <- run_rf_analysis(
      df           = df_cl,
      response_var = "Abundance",
      predictors   = c(litter_vars, clim_vars_MATMAP),
      ntree        = 1000, cv_folds = 10, nperm_oob = 500,
      nrep_vi      = 500,  n_boot = 100, grid.size = 20)
    res$summary_table$Clade  <- clade
    res$pdp_ci_df$Clade      <- clade
    res$imp_results_rf$Clade <- clade
    res$rf_cv$results$Clade  <- clade
    res_list[[clade]] <- res
  }
  res_list
})

all_summaries_MATMAP <- bind_rows(lapply(clade_results_MATMAP, `[[`, "summary_table"))
all_pdp_ci_MATMAP    <- bind_rows(lapply(clade_results_MATMAP, `[[`, "pdp_ci_df"))
all_imp_MATMAP       <- bind_rows(lapply(clade_results_MATMAP, `[[`, "imp_results_rf"))

sig_lookup_MATMAP <- map_dfr(names(clade_results_MATMAP), function(cl) {
  tibble(Clade = cl, feature = clade_results_MATMAP[[cl]]$sig_vars_rfp)
})

write.table(all_summaries_MATMAP,
            file = here::here("plots/RF_validation_clades_MATMAP.csv"),
            quote = F, row.names = F, col.names = T, sep = ",")

all_imp_sig_MATMAP <- all_imp_MATMAP %>%
  semi_join(sig_lookup_MATMAP, by = c("Clade", "feature")) %>%
  mutate(
    feature = gsub("_", " ", feature),
    feature = factor(feature, levels = feature_levels_MATMAP)
  )

pos_df_clade_MATMAP <- all_imp_sig_MATMAP %>%
  group_by(Clade) %>%
  reframe(x = max(importance, na.rm = TRUE) * 1.05, y = n() + 0.3)

annot_df_clade_MATMAP <- all_summaries_MATMAP %>%
  filter(Metric %in% c("10-fold CV R²", "Permuted CV p-value")) %>%
  dplyr::select(Clade, Metric, Value) %>%
  pivot_wider(names_from = Metric, values_from = Value) %>%
  dplyr::rename(CV_R2 = `10-fold CV R²`, P_val = `Permuted CV p-value`) %>%
  mutate(
    CV_R2 = as.numeric(CV_R2),
    P_num = as.numeric(sub("^<", "", P_val)),
    expr  = paste0("atop(CV~R^2==", round(CV_R2, 3), ",italic(P)<",
                   format.pval(P_num, digits = 2, eps = 1e-3), ")")
  ) %>%
  left_join(pos_df_clade_MATMAP, by = "Clade") %>%
  dplyr::filter(!Clade %in% c("V", "UND"))

imp_plot_rf_clades_MATMAP <- all_imp_sig_MATMAP %>%
  mutate(var_type = case_when(feature %in% clim_vars_MATMAP ~ "Climate",
                              feature %in% gsub("_", " ", litter_vars) ~ "Litter chemistry",
                              TRUE ~ "All"),
         Clade = factor(Clade, levels = c("I", "II", "III", "IV", "V", "UND"))) |>
  dplyr::filter(Clade != "UND") |>
  ggplot(aes(x = importance, y = feature, color = feature)) +
  scale_color_manual(values = allvars_pallete[feature_levels_MATMAP],
                     guide = guide_legend(ncol = 2)) +
  scale_y_discrete(limits = rev(feature_levels_MATMAP)) +
  scale_x_continuous(labels = scales::percent_format()) +
  ggh4x::facet_grid2(Clade ~ ., scales = "free_y", space = "free_y",
                     strip = ggh4x::strip_themed(
                       background_y = ggh4x::elem_list_rect(fill = clade_colors[-5])
                     )) +
  geom_point(position = position_dodge(width = 0.4), size = 2.5) +
  geom_errorbarh(aes(xmin = importance.05, xmax = importance.95),
                 position = position_dodge(width = 0.4), height = 0, linewidth = 1.5) +
  geom_text(data = annot_df_clade_MATMAP,
            aes(x = 0.03, y = 1, label = expr),
            parse = TRUE, inherit.aes = FALSE, hjust = 0, vjust = 0, size = 3) +
  apatheme +
  theme(axis.text.y = element_blank(), axis.title.y = element_blank(),
        axis.line.y = element_blank(), strip.text = element_text(size = 10),
        panel.grid.major.x = element_line(colour = "grey90"),
        panel.border = element_rect(colour = "grey90", fill = NA, linewidth = 0.5),
        legend.spacing = unit(0.1, "cm"), legend.key.size = unit(0.3, "cm"),
        legend.margin = ggplot2::margin(0, 0, 0, 0),
        legend.text = element_text(size = 9), legend.position = "bottom")

rug_clades_MATMAP <- reshape2::melt(clade_results_MATMAP$I$rug) %>%
  dplyr::rename(feature = variable) |>
  mutate(var_type = case_when(feature %in% clim_vars_MATMAP ~ "Climate",
                              feature %in% litter_vars       ~ "Litter chemistry",
                              TRUE ~ NA_character_),
         feature  = gsub("_", " ", feature),
         feature  = factor(feature, levels = feature_levels_MATMAP))

pdp_plot_clades_MATMAP <- all_pdp_ci_MATMAP %>%
  mutate(var_type = case_when(feature %in% clim_vars_MATMAP ~ "Climate",
                              feature %in% litter_vars       ~ "Litter chemistry",
                              TRUE ~ NA_character_)) |>
  mutate(var_type = factor(var_type, levels = c("Climate", "Litter chemistry")),
         feature  = gsub("_", " ", feature),
         feature  = factor(feature, levels = feature_levels_MATMAP)) |>
  dplyr::filter(!Clade %in% c("UND", "V")) |>
  ggplot(aes(x = value, y = mean_pred, color = Clade, linetype = significant)) +
  geom_line(size = 1) +
  scale_y_continuous(labels = percent_format(scale = 100)) +
  scale_color_manual(values = clade_colors, limits = c("I", "II", "III", "IV", "UND")) +
  scale_linetype_manual(name = NULL,
                        values = c("TRUE" = "solid", "FALSE" = "longdash"),
                        labels = c(expression(P > 0.05), expression(P < 0.05))) +
  geom_rug(data = rug_clades_MATMAP, aes(x = value), sides = "b", inherit.aes = FALSE) +
  ggh4x::facet_nested(~ var_type + feature, scales = "free_x") +
  labs(x = "Variable value", y = "Predicted relative abundance") +
  guides(color = "none", fill = "none",
         linetype = guide_legend(order = 1, override.aes = list(size = 1))) +
  apatheme +
  theme(axis.title.y = element_text(size = 8, margin = ggplot2::margin(r = 10)),
        axis.text.y  = element_text(size = 8),
        strip.background = element_rect(fill = NA, colour = "black"),
        strip.text   = element_text(size = 10),
        panel.grid.major.y = element_line(colour = "grey90"),
        text = element_text(size = 10),
        legend.position = c(0.02, 0.98), legend.justification = c(0, 1),
        legend.background = element_rect(fill = alpha("white", 0.7), colour = "black"),
        legend.key.size = unit(0.3, "cm"), legend.text = element_text(size = 6),
        legend.key.width = unit(1, "cm"), legend.key.height = unit(0.5, "cm"),
        legend.title = element_text(size = 8))

print(pdp_plot_clades_MATMAP)
print(imp_plot_rf_clades_MATMAP)

ggarrange(pdp_plot_clades_MATMAP, imp_plot_rf_clades_MATMAP,
          labels = "AUTO", ncol = 2, common.legend = F,
          legend = "bottom", widths = c(1.5, 0.5), hjust = -0.05, vjust = 1)

ggsave(here::here("plots/plots_pdp/pdp_plot_rf_clades_MAT_MAP_litt_vars.png"),
       width = width_mm, height = width_mm * 0.5, units = "mm", dpi = 300)
```

##### Subclade level — RF (MAT & MAP)

```{r pdp-subclades-rf-mat-map, message=FALSE, warning=FALSE, results='hold', echo=FALSE, fig.width=25, fig.height=20}
set.seed(711)

subclade_results_MATMAP <- cached(here::here("cache/subclade_results_MATMAP.rds"), {
  res_list <- list()
  for (subclade in unique(subclades_Ab_md$Subclade)) {
    message("Processing subclade: ", subclade)
    df_cl <- subclades_Ab_md %>%
      filter(Subclade == subclade) %>%
      dplyr::select(all_of(c("Abundance", litter_vars, clim_vars_MATMAP))) %>%
      na.omit()
    res <- run_rf_analysis(
      df           = df_cl,
      response_var = "Abundance",
      predictors   = c(litter_vars, clim_vars_MATMAP),
      ntree        = 1000, cv_folds = 10, nperm_oob = 500,
      nrep_vi      = 100,  n_boot = 100, grid.size = 20)
    res$summary_table$Subclade  <- subclade
    res$pdp_ci_df$Subclade      <- subclade
    res$imp_results_rf$Subclade <- subclade
    res$rf_cv$results$Subclade  <- subclade
    res_list[[subclade]] <- res
  }
  res_list
})

all_summaries_sub_MATMAP <- bind_rows(lapply(subclade_results_MATMAP, `[[`, "summary_table"))
all_pdp_ci_sub_MATMAP    <- bind_rows(lapply(subclade_results_MATMAP, `[[`, "pdp_ci_df"))
all_imp_sub_MATMAP       <- bind_rows(lapply(subclade_results_MATMAP, `[[`, "imp_results_rf"))

sig_lookup_sub_MATMAP <- map_dfr(names(subclade_results_MATMAP), function(cl) {
  tibble(Subclade = cl, feature = subclade_results_MATMAP[[cl]]$sig_vars_rfp)
})

write.table(all_summaries_sub_MATMAP,
            file = here::here("plots/RF_validation_subclades_MATMAP.csv"),
            quote = F, row.names = F, col.names = T, sep = ",")

all_imp_sig_sub_MATMAP <- all_imp_sub_MATMAP %>%
  semi_join(sig_lookup_sub_MATMAP, by = c("Subclade", "feature")) %>%
  mutate(
    feature = gsub("_", " ", feature),
    feature = factor(feature, levels = feature_levels_MATMAP)
  )

# --- Annotation (high-abundance subclades) ---

pos_df_sub_MATMAP <- all_imp_sig_sub_MATMAP %>%
  dplyr::filter(Subclade %in% c("IC", "IIG", "IIIA", "IVB")) |>
  group_by(Subclade) %>%
  reframe(x = max(importance, na.rm = TRUE) * 1.05, y = n() + 0.3)

annot_df_sub_MATMAP <- all_summaries_sub_MATMAP %>%
  filter(Metric %in% c("10-fold CV R²", "Permuted CV p-value")) %>%
  dplyr::select(Subclade, Metric, Value) %>%
  pivot_wider(names_from = Metric, values_from = Value) %>%
  dplyr::rename(CV_R2 = `10-fold CV R²`, P_val = `Permuted CV p-value`) %>%
  mutate(
    CV_R2 = as.numeric(CV_R2),
    P_num = as.numeric(sub("^<", "", P_val)),
    expr  = paste0("atop(CV~R^2==", round(CV_R2, 3), ",italic(P)<",
                   format.pval(P_num, digits = 2, eps = 1e-3), ")")
  ) %>%
  left_join(pos_df_sub_MATMAP, by = "Subclade") %>%
  dplyr::filter(Subclade %in% c("IC", "IIG", "IIIA", "IVB")) %>%
  mutate(Subclade = factor(Subclade, levels = c("IIIA", "IC", "IVB", "IIG")))

rug_sub_MATMAP <- reshape2::melt(subclade_results_MATMAP$IIIA$rug) %>%
  dplyr::rename(feature = variable) |>
  mutate(var_type = case_when(feature %in% clim_vars_MATMAP ~ "Climate",
                              feature %in% litter_vars       ~ "Litter chemistry",
                              TRUE ~ NA_character_),
         feature  = gsub("_", " ", feature),
         feature  = factor(feature, levels = feature_levels_MATMAP))

# --- High-abundance subclades ---

imp_plot_rf_sub_MATMAP <- all_imp_sig_sub_MATMAP %>%
  dplyr::filter(Subclade %in% c("IC", "IIG", "IIIA", "IVB")) |>
  mutate(Subclade = factor(Subclade, levels = c("IIIA", "IC", "IVB", "IIG"))) |>
  ggplot(aes(x = importance, y = feature, color = feature)) +
  scale_color_manual(values = allvars_pallete[feature_levels_MATMAP],
                     guide = guide_legend(ncol = 2)) +
  scale_y_discrete(limits = rev(feature_levels_MATMAP)) +
  scale_x_continuous(labels = scales::percent_format()) +
  ggh4x::facet_grid2(Subclade ~ ., scales = "free_y", space = "free_y",
                     strip = ggh4x::strip_themed(
                       background_y = ggh4x::elem_list_rect(
                         fill = subclade_colros[names(subclade_colros) %in%
                                                  c("IIIA", "IC", "IVB", "IIG")][c(3, 1, 4, 2)]
                       )
                     )) +
  geom_point(position = position_dodge(width = 0.6), size = 3) +
  geom_errorbarh(aes(xmin = importance.05, xmax = importance.95),
                 position = position_dodge(width = 0.6), height = 0, linewidth = 1.5) +
  geom_text(data = annot_df_sub_MATMAP |>
              dplyr::filter(Subclade %in% c("IIIA", "IC", "IVB", "IIG")),
            aes(x = 0.0275, y = 1, label = expr),
            parse = TRUE, inherit.aes = FALSE, hjust = 0, vjust = 0, size = 3) +
  xlab("importance (RMSE)") +
  apatheme +
  theme(axis.text.y = element_blank(), axis.title.y = element_blank(),
        axis.line.y = element_blank(), strip.text = element_text(size = 10),
        panel.grid.major.x = element_line(colour = "grey90"),
        panel.border = element_rect(colour = "grey90", fill = NA, linewidth = 0.5),
        legend.spacing = unit(0.1, "cm"), legend.key.size = unit(0.3, "cm"),
        legend.margin = ggplot2::margin(0, 0, 0, 0),
        legend.text = element_text(size = 9), legend.position = "bottom")

pdp_plot_sub_MATMAP <- all_pdp_ci_sub_MATMAP %>%
  mutate(var_type = case_when(feature %in% clim_vars_MATMAP ~ "Climate",
                              feature %in% litter_vars       ~ "Litter chemistry",
                              TRUE ~ NA_character_)) |>
  mutate(var_type = factor(var_type, levels = c("Climate", "Litter chemistry")),
         feature  = gsub("_", " ", feature),
         feature  = factor(feature, levels = feature_levels_MATMAP)) |>
  dplyr::filter(Subclade %in% c("IC", "IIG", "IIIA", "IVB")) |>
  ggplot(aes(x = value, y = mean_pred, color = Subclade, linetype = significant)) +
  geom_line(size = 1) +
  scale_y_continuous(labels = percent_format(scale = 100)) +
  scale_color_manual(values = subclade_colors_bp, limits = c("IC", "IIG", "IIIA", "IVB")) +
  scale_linetype_manual(name = NULL,
                        values = c("TRUE" = "solid", "FALSE" = "longdash"),
                        labels = c(expression(P > 0.05), expression(P < 0.05))) +
  geom_rug(data = rug_sub_MATMAP, aes(x = value), sides = "b", inherit.aes = FALSE) +
  ggh4x::facet_nested(~ var_type + feature, scales = "free_x") +
  labs(x = "Variable value", y = "Predicted relative abundance") +
  guides(color = "none", fill = "none",
         linetype = guide_legend(order = 1, override.aes = list(size = 1))) +
  apatheme +
  theme(axis.title.y = element_text(size = 8, margin = ggplot2::margin(r = 10)),
        axis.text.y = element_text(size = 8), axis.text.x = element_text(size = 8),
        strip.background = element_rect(fill = NA, colour = "black"),
        strip.text = element_text(size = 10),
        panel.grid.major.y = element_line(colour = "grey90"),
        text = element_text(size = 10),
        legend.position = c(0.02, 0.98), legend.justification = c(0, 1),
        legend.background = element_rect(fill = alpha("white", 0.7), colour = "black"),
        legend.key.size = unit(0.3, "cm"), legend.text = element_text(size = 6),
        legend.key.width = unit(1, "cm"), legend.key.height = unit(0.5, "cm"),
        legend.title = element_text(size = 8))

print(pdp_plot_sub_MATMAP)

imp_plot_rf_sub_MATMAP$layers[[3]] <- NULL  # remove annotation for combined plot
ggarrange(pdp_plot_sub_MATMAP, imp_plot_rf_sub_MATMAP,
          labels = "AUTO", ncol = 2, common.legend = F,
          legend = "bottom", widths = c(1.5, 0.5), hjust = -0.05, vjust = 1)

ggsave(here::here("plots/plots_pdp/pdp_plot_rf_subclades_MAT_MAP_litt_vars.png"),
       width = width_mm, height = width_mm * 0.5, units = "mm", dpi = 300, bg = "white")

# --- Low-abundance subclades ---

subs_lowAb_MATMAP <- intersect(
  correct_order[-c(1, 4, 5, 12, 16, 17, 23, 25, 29)],
  unique(all_imp_sig_sub_MATMAP$Subclade[
    !all_imp_sig_sub_MATMAP$Subclade %in% c("IC", "IIG", "IIIA", "IVB")])
)

imp_plot_rf_sub_MATMAP_lowAb <- all_imp_sig_sub_MATMAP %>%
  dplyr::filter(!Subclade %in% c("IC", "IIG", "IIIA", "IVB")) |>
  mutate(Subclade = factor(Subclade, levels = subs_lowAb_MATMAP)) |>
  ggplot(aes(x = importance, y = feature, color = feature)) +
  scale_color_manual(values = allvars_pallete[feature_levels_MATMAP],
                     guide = guide_legend(ncol = 2)) +
  scale_y_discrete(limits = rev(feature_levels_MATMAP)) +
  scale_x_continuous(labels = scales::percent_format()) +
  ggh4x::facet_grid2(Subclade ~ ., scales = "free_y", space = "free_y",
                     strip = ggh4x::strip_themed(
                       background_y = ggh4x::elem_list_rect(
                         fill = subclade_colors_bp[subs_lowAb_MATMAP]
                       )
                     )) +
  geom_point(position = position_dodge(width = 0.6), size = 3) +
  geom_errorbarh(aes(xmin = importance.05, xmax = importance.95),
                 position = position_dodge(width = 0.6), height = 0, linewidth = 1.5) +
  xlab("importance (RMSE)") +
  apatheme +
  theme(axis.text.y = element_blank(), axis.title.y = element_blank(),
        axis.line.y = element_blank(), strip.text = element_text(size = 10),
        panel.grid.major.x = element_line(colour = "grey90"),
        panel.border = element_rect(colour = "grey90", fill = NA, linewidth = 0.5),
        legend.spacing = unit(0.1, "cm"), legend.key.size = unit(0.3, "cm"),
        legend.margin = ggplot2::margin(0, 0, 0, 0),
        legend.text = element_text(size = 9))

pdp_plot_sub_MATMAP_lowAb <- all_pdp_ci_sub_MATMAP %>%
  mutate(Clade = case_when(Subclade %in% clade_I_subclades   ~ "I",
                           Subclade %in% clade_II_subclades  ~ "II",
                           Subclade %in% clade_III_subclades ~ "III",
                           Subclade %in% clade_IV_subclades  ~ "IV",
                           Subclade %in% clade_V_subclades   ~ "V")) %>%
  mutate(var_type = case_when(feature %in% clim_vars_MATMAP ~ "Climate",
                              feature %in% litter_vars       ~ "Litter chemistry",
                              TRUE ~ NA_character_)) |>
  mutate(var_type = factor(var_type, levels = c("Climate", "Litter chemistry")),
         feature  = gsub("_", " ", feature),
         feature  = factor(feature, levels = feature_levels_MATMAP)) |>
  dplyr::filter(!Subclade %in% c("IC", "IIG", "IIIA", "IVB")) |>
  ggplot(aes(x = value, y = mean_pred, color = Subclade, linetype = significant)) +
  geom_line(size = 1) +
  scale_y_continuous(labels = percent_format(scale = 100)) +
  scale_color_manual(values = subclade_colors_bp) +
  scale_linetype_manual(name = NULL,
                        values = c("TRUE" = "solid", "FALSE" = "longdash"),
                        labels = c(expression(P > 0.05), expression(P < 0.05))) +
  geom_rug(data = rug_sub_MATMAP, aes(x = value), sides = "b", inherit.aes = FALSE) +
  ggh4x::facet_nested(Clade ~ var_type + feature, scales = "free_x") +
  labs(x = "Variable value", y = "Predicted relative abundance") +
  guides(color = "none", fill = "none",
         linetype = guide_legend(order = 1, override.aes = list(size = 1))) +
  apatheme +
  theme(axis.title.y = element_text(size = 8, margin = ggplot2::margin(r = 10)),
        axis.text.y = element_text(size = 8),
        strip.background = element_rect(fill = NA, colour = "black"),
        strip.text = element_text(size = 10),
        panel.grid.major.y = element_line(colour = "grey90"),
        text = element_text(size = 10),
        legend.position = c(0.02, 0.98), legend.justification = c(0, 1),
        legend.background = element_rect(fill = alpha("white", 0.7), colour = "black"),
        legend.key.size = unit(0.3, "cm"), legend.text = element_text(size = 6),
        legend.key.width = unit(1, "cm"), legend.key.height = unit(0.5, "cm"),
        legend.title = element_text(size = 8))

print(pdp_plot_sub_MATMAP_lowAb)

ggarrange(pdp_plot_sub_MATMAP_lowAb, imp_plot_rf_sub_MATMAP_lowAb,
          labels = "AUTO", ncol = 2, common.legend = F,
          legend = "bottom", widths = c(1.5, 0.5), hjust = -0.05, vjust = 1)

ggsave(here::here("plots/plots_pdp/pdp_plot_rf_subclades_MAT_MAP_litt_vars_lowAb.png"),
       width = width_mm, height = width_mm * 0.75, units = "mm", dpi = 300)
```

##### 2D heatmap PDPs — top 4 subclades, MAT & MAP (RF)

```{r pdp-subclades-rf-heatmap-mat-map, message=FALSE, warning=FALSE, results='hold', echo=FALSE, fig.width=20, fig.height=12}
subclades_2d_pdp_rf_MATMAP <- cached(here::here("cache/subclades_2d_pdp_rf_MATMAP.rds"), {
  map_dfr(top4_subclades, function(sub) {
    message("Subclade: ", sub)
    mod  <- subclade_results_MATMAP[[sub]]$rf_full
    X    <- subclade_results_MATMAP[[sub]]$rug
    y    <- subclade_results_MATMAP[[sub]]$rf_full$y
    pred <- Predictor$new(mod, data = X, y = y)

    map_dfr(clim_vars_MATMAP, function(cv) {
      map_dfr(litter_vars, function(lv) {
        message("  ", cv, " x ", lv)
        eff <- FeatureEffect$new(pred, feature = c(cv, lv), method = "pdp", grid.size = 20)
        eff$results |>
          dplyr::rename(clim_value = all_of(cv), litter_value = all_of(lv)) |>
          dplyr::mutate(clim_var = cv, litter_var = lv, Subclade = sub)
      })
    })
  })
})

heatmap_2d_rf_MATMAP <- function(sub) {
  subclades_2d_pdp_rf_MATMAP |>
    dplyr::filter(Subclade == sub) |>
    mutate(
      litter_var = gsub("_", " ", litter_var),
      litter_var = factor(litter_var, levels = gsub("_", " ", litter_vars)),
      clim_var   = factor(clim_var,   levels = clim_vars_MATMAP)
    ) |>
    ggplot(aes(x = litter_value, y = clim_value, fill = .value)) +
    geom_raster(interpolate = TRUE) +
    scale_fill_viridis_c(name = "Predicted\nabundance",
                         labels = scales::percent_format(scale = 100)) +
    facet_grid(clim_var ~ litter_var, scales = "free") +
    labs(x = "Litter chemistry variable value",
         y = "Climate variable value", title = sub) +
    apatheme +
    theme(plot.title        = element_text(hjust = 0.5, face = "bold", size = 11,
                                           colour = subclade_colors_bp[sub]),
          strip.background  = element_rect(fill = NA, colour = "black"),
          strip.text        = element_text(size = 8),
          axis.text         = element_text(size = 7),
          axis.title        = element_text(size = 9),
          panel.spacing     = unit(0.3, "cm"),
          legend.key.height = unit(1, "cm"))
}

ggarrange(plotlist = lapply(top4_subclades, heatmap_2d_rf_MATMAP),
          ncol = 2, nrow = 2, common.legend = TRUE, legend = "right")

ggsave(here::here("plots/plots_pdp/pdp_heatmap_rf_top4_subclades_MATMAP.png"),
       width = 300, height = 200, units = "mm", dpi = 300, bg = "white")
```

##### 2D heatmap PDPs — significant predictors only, MAT & MAP (RF)

```{r pdp-subclades-rf-heatmap-mat-map-sig, message=FALSE, warning=FALSE, results='hold', echo=FALSE, fig.width=16, fig.height=10}
sig_pairs_MATMAP <- lapply(setNames(top4_subclades, top4_subclades), function(sub) {
  sig      <- subclade_results_MATMAP[[sub]]$sig_vars_rfp
  sig_clim <- intersect(sig, clim_vars_MATMAP)
  sig_litt <- intersect(sig, litter_vars)
  expand.grid(clim_var = sig_clim, litter_var = sig_litt, stringsAsFactors = FALSE)
})
sig_pairs_MATMAP <- Filter(function(x) nrow(x) > 0, sig_pairs_MATMAP)

message("Subclades with significant MAT/MAP × litter pairs: ",
        paste(names(sig_pairs_MATMAP), collapse = ", "))

subclades_2d_pdp_rf_MATMAP_sig <- cached(here::here("cache/subclades_2d_pdp_rf_MATMAP_sig.rds"), {
  map_dfr(names(sig_pairs_MATMAP), function(sub) {
    message("Subclade: ", sub)
    mod  <- subclade_results_MATMAP[[sub]]$rf_full
    X    <- subclade_results_MATMAP[[sub]]$rug
    y    <- subclade_results_MATMAP[[sub]]$rf_full$y
    pred <- Predictor$new(mod, data = X, y = y)

    pmap_dfr(sig_pairs_MATMAP[[sub]], function(clim_var, litter_var) {
      message("  ", clim_var, " x ", litter_var)
      eff <- FeatureEffect$new(pred, feature = c(clim_var, litter_var),
                               method = "pdp", grid.size = 20)
      eff$results |>
        dplyr::rename(clim_value = all_of(clim_var), litter_value = all_of(litter_var)) |>
        dplyr::mutate(clim_var = clim_var, litter_var = litter_var, Subclade = sub)
    })
  })
})

heatmap_2d_rf_MATMAP_sig <- function(sub) {
  subclades_2d_pdp_rf_MATMAP_sig |>
    dplyr::filter(Subclade == sub) |>
    mutate(
      litter_var = gsub("_", " ", litter_var),
      litter_var = factor(litter_var, levels = gsub("_", " ", litter_vars)),
      clim_var   = factor(clim_var,   levels = clim_vars_MATMAP)
    ) |>
    ggplot(aes(x = litter_value, y = clim_value, fill = .value)) +
    geom_raster(interpolate = TRUE) +
    scale_fill_viridis_c(name = "Predicted\nabundance",
                         labels = scales::percent_format(scale = 100),
                         limits = c(0, 0.60), oob = scales::squish) +
    facet_grid(clim_var ~ litter_var, scales = "free") +
    labs(x = "Litter chemistry variable value",
         y = "Climate variable value", title = sub) +
    apatheme +
    theme(plot.title        = element_text(hjust = 0.5, face = "bold", size = 11,
                                           colour = subclade_colors_bp[sub]),
          strip.background  = element_rect(fill = NA, colour = "black"),
          strip.text        = element_text(size = 8),
          axis.text         = element_text(size = 7),
          axis.title        = element_text(size = 9),
          panel.spacing     = unit(0.3, "cm"),
          legend.key.height = unit(1, "cm"))
}

pdp_heatmaps_rf_MATMAP_sig <- lapply(names(sig_pairs_MATMAP), heatmap_2d_rf_MATMAP_sig)

ggarrange(plotlist = pdp_heatmaps_rf_MATMAP_sig,
          ncol = length(pdp_heatmaps_rf_MATMAP_sig), nrow = 1,
          common.legend = TRUE, legend = "right")

ggsave(here::here("plots/plots_pdp/pdp_heatmap_rf_top4_subclades_MATMAP_sig_vars.png"),
       width = 100 * length(pdp_heatmaps_rf_MATMAP_sig), height = 160,
       units = "mm", dpi = 300, bg = "white")
```

##### 2D heatmap PDPs — top 4 subclades (GLM)

```{r pdp-subclades-glm-heatmap, message=FALSE, warning=FALSE, results='hold', echo=FALSE, fig.width=20, fig.height=12}
top4_subclades <- c("IIIA", "IC", "IVB", "IIG")

# Compute 2D PDPs for every climate × litter combination for each of the top 4 subclades
subclades_2d_pdp <- cached(here::here("cache/subclades_2d_pdp.rds"), {
  map_dfr(top4_subclades, function(sub) {
    message("Subclade: ", sub)
    mod  <- subclade_glm_results[[sub]]$glm_full
    X    <- subclade_glm_results[[sub]]$rug
    y    <- subclade_glm_results[[sub]]$glm_full$y
    pred <- Predictor$new(mod, data = X, y = y)

    map_dfr(clim_vars_PC, function(cv) {
      map_dfr(litter_vars, function(lv) {
        message("  ", cv, " x ", lv)
        eff <- FeatureEffect$new(pred, feature = c(cv, lv), method = "pdp", grid.size = 20)
        eff$results |>
          dplyr::rename(clim_value = all_of(cv), litter_value = all_of(lv)) |>
          dplyr::mutate(clim_var = cv, litter_var = lv, Subclade = sub)
      })
    })
  })
})

# Helper: one heatmap faceted by climate var (rows) × litter var (columns)
heatmap_2d_pdp <- function(sub) {
  subclades_2d_pdp |>
    dplyr::filter(Subclade == sub) |>
    mutate(
      clim_var   = dplyr::recode(gsub("[._]", " ", clim_var),
                                  `PC1 clim` = "PC1", `PC2 clim` = "PC2"),
      litter_var = gsub("_", " ", litter_var),
      litter_var = factor(litter_var, levels = gsub("_", " ", litter_vars)),
      clim_var   = factor(clim_var,   levels = c("PC1", "PC2"))
    ) |>
    ggplot(aes(x = litter_value, y = clim_value, fill = .value)) +
    geom_raster(interpolate = TRUE) +
    scale_fill_viridis_c(name = "Predicted\nabundance",
                         labels = scales::percent_format(scale = 100)) +
    facet_grid(clim_var ~ litter_var, scales = "free") +
    labs(x = "Litter chemistry variable value",
         y = "Climate variable value",
         title = sub) +
    apatheme +
    theme(plot.title       = element_text(hjust = 0.5, face = "bold", size = 11,
                                          colour = subclade_colors_bp[sub]),
          strip.background = element_rect(fill = NA, colour = "black"),
          strip.text       = element_text(size = 8),
          axis.text        = element_text(size = 7),
          axis.title       = element_text(size = 9),
          panel.spacing    = unit(0.3, "cm"),
          legend.key.height = unit(1, "cm"))
}

pdp_heatmaps <- lapply(top4_subclades, heatmap_2d_pdp)

ggarrange(plotlist = pdp_heatmaps, ncol = 2, nrow = 2,
          common.legend = TRUE, legend = "right")

ggsave(here::here("plots/plots_pdp/pdp_heatmap_glm_top4_subclades.png"),
       width = 300, height = 200, units = "mm", dpi = 300, bg = "white")
```

#### GLM + PDP with MAT & MAP as climate variables

##### Clade level — GLM (MAT & MAP)

```{r pdp-clades-glm-mat-map, message=FALSE, warning=FALSE, results='hold', echo=FALSE, fig.width=25, fig.height=20}
set.seed(711)

clade_glm_results_MATMAP <- cached(here::here("cache/clade_glm_results_MATMAP.rds"), {
  res_list <- list()
  for (clade in unique(clades_Ab_md$Clade)) {
    message("Processing clade: ", clade)
    df_cl <- clades_Ab_md %>%
      filter(Clade == clade) %>%
      dplyr::select(all_of(c("Abundance", litter_vars, clim_vars_MATMAP))) %>%
      na.omit()
    res <- run_glm_analysis(
      df           = df_cl,
      response_var = "Abundance",
      predictors   = c(litter_vars, clim_vars_MATMAP),
      cv_folds = 10, nrep_vi = 500, n_boot = 100, grid.size = 20)
    res$summary_table$Clade  <- clade
    res$pdp_ci_df$Clade      <- clade
    res$imp_results_rf$Clade <- clade
    res_list[[clade]] <- res
  }
  res_list
})

all_summaries_glm_MATMAP <- bind_rows(lapply(clade_glm_results_MATMAP, `[[`, "summary_table"))
all_pdp_ci_glm_MATMAP    <- bind_rows(lapply(clade_glm_results_MATMAP, `[[`, "pdp_ci_df"))
all_imp_glm_MATMAP       <- bind_rows(lapply(clade_glm_results_MATMAP, `[[`, "imp_results_rf"))

sig_lookup_glm_MATMAP <- map_dfr(names(clade_glm_results_MATMAP), function(cl) {
  tibble(Clade = cl, feature = clade_glm_results_MATMAP[[cl]]$sig_vars_rfp)
})

write.table(all_summaries_glm_MATMAP,
            file = here::here("plots/GLM_validation_clades_MATMAP.csv"),
            quote = F, row.names = F, col.names = T, sep = ",")

all_imp_sig_glm_MATMAP <- all_imp_glm_MATMAP %>%
  semi_join(sig_lookup_glm_MATMAP, by = c("Clade", "feature")) %>%
  mutate(feature = gsub("_", " ", feature),
         feature = factor(feature, levels = feature_levels_MATMAP))

pos_df_clade_glm_MATMAP <- all_imp_sig_glm_MATMAP %>%
  group_by(Clade) %>%
  reframe(x = max(importance, na.rm = TRUE) * 1.05, y = n() + 0.3)

annot_df_clade_glm_MATMAP <- all_summaries_glm_MATMAP %>%
  filter(Metric %in% c("10-fold CV R²", "Overall model p-value (F-test)")) %>%
  dplyr::select(Clade, Metric, Value) %>%
  pivot_wider(names_from = Metric, values_from = Value) %>%
  dplyr::rename(CV_R2 = `10-fold CV R²`, P_val = `Overall model p-value (F-test)`) %>%
  mutate(
    CV_R2 = as.numeric(CV_R2),
    P_num = as.numeric(sub("^<", "", P_val)),
    expr  = paste0("atop(CV~R^2==", round(CV_R2, 3), ",italic(P)==",
                   format.pval(P_num, digits = 2, eps = 1e-3), ")")
  ) %>%
  left_join(pos_df_clade_glm_MATMAP, by = "Clade") %>%
  dplyr::filter(!Clade %in% c("V", "UND"))

imp_plot_glm_clades_MATMAP <- all_imp_sig_glm_MATMAP %>%
  mutate(var_type = case_when(feature %in% clim_vars_MATMAP ~ "Climate",
                              feature %in% gsub("_", " ", litter_vars) ~ "Litter chemistry",
                              TRUE ~ "All"),
         Clade = factor(Clade, levels = c("I", "II", "III", "IV", "V", "UND"))) |>
  dplyr::filter(Clade != "UND") |>
  ggplot(aes(x = importance, y = feature, color = feature)) +
  scale_color_manual(values = allvars_pallete[feature_levels_MATMAP],
                     guide = guide_legend(ncol = 2)) +
  scale_y_discrete(limits = rev(feature_levels_MATMAP)) +
  scale_x_continuous(labels = scales::percent_format()) +
  ggh4x::facet_grid2(Clade ~ ., scales = "free_y", space = "free_y",
                     strip = ggh4x::strip_themed(
                       background_y = ggh4x::elem_list_rect(fill = clade_colors[-5])
                     )) +
  geom_point(position = position_dodge(width = 0.4), size = 2.5) +
  geom_errorbarh(aes(xmin = importance.05, xmax = importance.95),
                 position = position_dodge(width = 0.4), height = 0, linewidth = 1.5) +
  geom_text(data = annot_df_clade_glm_MATMAP,
            aes(x = 0.03, y = 1, label = expr),
            parse = TRUE, inherit.aes = FALSE, hjust = 0, vjust = 0, size = 3) +
  apatheme +
  theme(axis.text.y = element_blank(), axis.title.y = element_blank(),
        axis.line.y = element_blank(), strip.text = element_text(size = 10),
        panel.grid.major.x = element_line(colour = "grey90"),
        panel.border = element_rect(colour = "grey90", fill = NA, linewidth = 0.5),
        legend.spacing = unit(0.1, "cm"), legend.key.size = unit(0.3, "cm"),
        legend.margin = ggplot2::margin(0, 0, 0, 0),
        legend.text = element_text(size = 9), legend.position = "bottom")

rug_clades_glm_MATMAP <- reshape2::melt(clade_glm_results_MATMAP$I$rug) %>%
  dplyr::rename(feature = variable) |>
  mutate(var_type = case_when(feature %in% clim_vars_MATMAP ~ "Climate",
                              feature %in% litter_vars       ~ "Litter chemistry",
                              TRUE ~ NA_character_),
         feature  = gsub("_", " ", feature),
         feature  = factor(feature, levels = feature_levels_MATMAP))

pdp_plot_clades_glm_MATMAP <- all_pdp_ci_glm_MATMAP %>%
  mutate(var_type = case_when(feature %in% clim_vars_MATMAP ~ "Climate",
                              feature %in% litter_vars       ~ "Litter chemistry",
                              TRUE ~ NA_character_)) |>
  mutate(var_type = factor(var_type, levels = c("Climate", "Litter chemistry")),
         feature  = gsub("_", " ", feature),
         feature  = factor(feature, levels = feature_levels_MATMAP)) |>
  dplyr::filter(!Clade %in% c("UND", "V")) |>
  ggplot(aes(x = value, y = mean_pred, color = Clade, linetype = significant)) +
  geom_line(size = 1) +
  scale_y_continuous(labels = percent_format(scale = 100)) +
  scale_color_manual(values = clade_colors, limits = c("I", "II", "III", "IV", "UND")) +
  scale_linetype_manual(name = NULL,
                        values = c("TRUE" = "solid", "FALSE" = "longdash"),
                        labels = c(expression(P > 0.05), expression(P < 0.05))) +
  geom_rug(data = rug_clades_glm_MATMAP, aes(x = value), sides = "b", inherit.aes = FALSE) +
  ggh4x::facet_nested(~ var_type + feature, scales = "free_x") +
  labs(x = "Variable value", y = "Predicted relative abundance") +
  guides(color = "none", fill = "none",
         linetype = guide_legend(order = 1, override.aes = list(size = 1))) +
  apatheme +
  theme(axis.title.y = element_text(size = 8, margin = ggplot2::margin(r = 10)),
        axis.text.y  = element_text(size = 8),
        strip.background = element_rect(fill = NA, colour = "black"),
        strip.text   = element_text(size = 10),
        panel.grid.major.y = element_line(colour = "grey90"),
        text = element_text(size = 10),
        legend.position = c(0.02, 0.98), legend.justification = c(0, 1),
        legend.background = element_rect(fill = alpha("white", 0.7), colour = "black"),
        legend.key.size = unit(0.3, "cm"), legend.text = element_text(size = 6),
        legend.key.width = unit(1, "cm"), legend.key.height = unit(0.5, "cm"),
        legend.title = element_text(size = 8))

print(pdp_plot_clades_glm_MATMAP)
print(imp_plot_glm_clades_MATMAP)

ggarrange(pdp_plot_clades_glm_MATMAP, imp_plot_glm_clades_MATMAP,
          labels = "AUTO", ncol = 2, common.legend = F,
          legend = "bottom", widths = c(1.5, 0.5), hjust = -0.05, vjust = 1)

ggsave(here::here("plots/plots_pdp/pdp_plot_glm_clades_MAT_MAP_litt_vars.png"),
       width = width_mm, height = width_mm * 0.5, units = "mm", dpi = 300)
```

##### Subclade level — GLM (MAT & MAP)

```{r pdp-subclades-glm-mat-map, message=FALSE, warning=FALSE, results='hold', echo=FALSE, fig.width=25, fig.height=20}
set.seed(711)

subclade_glm_results_MATMAP <- cached(here::here("cache/subclade_glm_results_MATMAP.rds"), {
  res_list <- list()
  for (subclade in unique(subclades_Ab_md$Subclade)) {
    message("Processing subclade: ", subclade)
    df_cl <- subclades_Ab_md %>%
      filter(Subclade == subclade) %>%
      dplyr::select(all_of(c("Abundance", litter_vars, clim_vars_MATMAP))) %>%
      na.omit()
    res <- run_glm_analysis(
      df           = df_cl,
      response_var = "Abundance",
      predictors   = c(litter_vars, clim_vars_MATMAP),
      cv_folds = 10, nrep_vi = 100, n_boot = 100, grid.size = 20)
    res$summary_table$Subclade  <- subclade
    res$pdp_ci_df$Subclade      <- subclade
    res$imp_results_rf$Subclade <- subclade
    res_list[[subclade]] <- res
  }
  res_list
})

all_summaries_glm_sub_MATMAP <- bind_rows(lapply(subclade_glm_results_MATMAP, `[[`, "summary_table"))
all_pdp_ci_glm_sub_MATMAP    <- bind_rows(lapply(subclade_glm_results_MATMAP, `[[`, "pdp_ci_df"))
all_imp_glm_sub_MATMAP       <- bind_rows(lapply(subclade_glm_results_MATMAP, `[[`, "imp_results_rf"))

sig_lookup_glm_sub_MATMAP <- map_dfr(names(subclade_glm_results_MATMAP), function(cl) {
  tibble(Subclade = cl, feature = subclade_glm_results_MATMAP[[cl]]$sig_vars_rfp)
})

write.table(all_summaries_glm_sub_MATMAP,
            file = here::here("plots/GLM_validation_subclades_MATMAP.csv"),
            quote = F, row.names = F, col.names = T, sep = ",")

all_imp_sig_glm_sub_MATMAP <- all_imp_glm_sub_MATMAP %>%
  semi_join(sig_lookup_glm_sub_MATMAP, by = c("Subclade", "feature")) %>%
  mutate(feature = gsub("_", " ", feature),
         feature = factor(feature, levels = feature_levels_MATMAP))

# --- Annotation (high-abundance subclades) ---

pos_df_glm_sub_MATMAP <- all_imp_sig_glm_sub_MATMAP %>%
  dplyr::filter(Subclade %in% c("IC", "IIG", "IIIA", "IVB")) |>
  group_by(Subclade) %>%
  reframe(x = max(importance, na.rm = TRUE) * 1.05, y = n() + 0.3)

annot_df_glm_sub_MATMAP <- all_summaries_glm_sub_MATMAP %>%
  filter(Metric %in% c("10-fold CV R²", "Overall model p-value (F-test)")) %>%
  dplyr::select(Subclade, Metric, Value) %>%
  pivot_wider(names_from = Metric, values_from = Value) %>%
  dplyr::rename(CV_R2 = `10-fold CV R²`, P_val = `Overall model p-value (F-test)`) %>%
  mutate(
    CV_R2 = as.numeric(CV_R2),
    P_num = as.numeric(sub("^<", "", P_val)),
    expr  = paste0("atop(CV~R^2==", round(CV_R2, 3), ",italic(P)==",
                   format.pval(P_num, digits = 2, eps = 1e-3), ")")
  ) %>%
  left_join(pos_df_glm_sub_MATMAP, by = "Subclade") %>%
  dplyr::filter(Subclade %in% c("IC", "IIG", "IIIA", "IVB")) %>%
  mutate(Subclade = factor(Subclade, levels = c("IIIA", "IC", "IVB", "IIG")))

rug_glm_sub_MATMAP <- reshape2::melt(subclade_glm_results_MATMAP$IIIA$rug) %>%
  dplyr::rename(feature = variable) |>
  mutate(var_type = case_when(feature %in% clim_vars_MATMAP ~ "Climate",
                              feature %in% litter_vars       ~ "Litter chemistry",
                              TRUE ~ NA_character_),
         feature  = gsub("_", " ", feature),
         feature  = factor(feature, levels = feature_levels_MATMAP))

# --- High-abundance subclades ---

imp_plot_glm_sub_MATMAP <- all_imp_sig_glm_sub_MATMAP %>%
  dplyr::filter(Subclade %in% c("IC", "IIG", "IIIA", "IVB")) |>
  mutate(Subclade = factor(Subclade, levels = c("IIIA", "IC", "IVB", "IIG"))) |>
  ggplot(aes(x = importance, y = feature, color = feature)) +
  scale_color_manual(values = allvars_pallete[feature_levels_MATMAP],
                     guide = guide_legend(ncol = 2)) +
  scale_y_discrete(limits = rev(feature_levels_MATMAP)) +
  scale_x_continuous(labels = scales::percent_format()) +
  ggh4x::facet_grid2(Subclade ~ ., scales = "free_y", space = "free_y",
                     strip = ggh4x::strip_themed(
                       background_y = ggh4x::elem_list_rect(
                         fill = subclade_colros[names(subclade_colros) %in%
                                                  c("IIIA", "IC", "IVB", "IIG")][c(4, 1, 3, 2)]
                       )
                     )) +
  geom_point(position = position_dodge(width = 0.6), size = 3) +
  geom_errorbarh(aes(xmin = importance.05, xmax = importance.95),
                 position = position_dodge(width = 0.6), height = 0, linewidth = 1.5) +
  geom_text(data = annot_df_glm_sub_MATMAP |>
              dplyr::filter(Subclade %in% c("IIIA", "IC", "IVB", "IIG")),
            aes(x = 0.0275, y = 1, label = expr),
            parse = TRUE, inherit.aes = FALSE, hjust = 0, vjust = 0, size = 3) +
  xlab("importance (RMSE)") +
  apatheme +
  theme(axis.text.y = element_blank(), axis.title.y = element_blank(),
        axis.line.y = element_blank(), strip.text = element_text(size = 10),
        panel.grid.major.x = element_line(colour = "grey90"),
        panel.border = element_rect(colour = "grey90", fill = NA, linewidth = 0.5),
        legend.spacing = unit(0.1, "cm"), legend.key.size = unit(0.3, "cm"),
        legend.margin = ggplot2::margin(0, 0, 0, 0),
        legend.text = element_text(size = 9), legend.position = "bottom")

pdp_plot_glm_sub_MATMAP <- all_pdp_ci_glm_sub_MATMAP %>%
  mutate(var_type = case_when(feature %in% clim_vars_MATMAP ~ "Climate",
                              feature %in% litter_vars       ~ "Litter chemistry",
                              TRUE ~ NA_character_)) |>
  mutate(var_type = factor(var_type, levels = c("Climate", "Litter chemistry")),
         feature  = gsub("_", " ", feature),
         feature  = factor(feature, levels = feature_levels_MATMAP)) |>
  dplyr::filter(Subclade %in% c("IC", "IIG", "IIIA", "IVB")) |>
  ggplot(aes(x = value, y = mean_pred, color = Subclade, linetype = significant)) +
  geom_line(size = 1) +
  scale_y_continuous(labels = percent_format(scale = 100)) +
  scale_color_manual(values = subclade_colors_bp, limits = c("IC", "IIG", "IIIA", "IVB")) +
  scale_linetype_manual(name = NULL,
                        values = c("TRUE" = "solid", "FALSE" = "longdash"),
                        labels = c(expression(P > 0.05), expression(P < 0.05))) +
  geom_rug(data = rug_glm_sub_MATMAP, aes(x = value), sides = "b", inherit.aes = FALSE) +
  ggh4x::facet_nested(~ var_type + feature, scales = "free_x") +
  labs(x = "Variable value", y = "Predicted relative abundance") +
  guides(color = "none", fill = "none",
         linetype = guide_legend(order = 1, override.aes = list(size = 1))) +
  apatheme +
  theme(axis.title.y = element_text(size = 8, margin = ggplot2::margin(r = 10)),
        axis.text.y = element_text(size = 8), axis.text.x = element_text(size = 8),
        strip.background = element_rect(fill = NA, colour = "black"),
        strip.text = element_text(size = 10),
        panel.grid.major.y = element_line(colour = "grey90"),
        text = element_text(size = 10),
        legend.position = c(0.02, 0.98), legend.justification = c(0, 1),
        legend.background = element_rect(fill = alpha("white", 0.7), colour = "black"),
        legend.key.size = unit(0.3, "cm"), legend.text = element_text(size = 6),
        legend.key.width = unit(1, "cm"), legend.key.height = unit(0.5, "cm"),
        legend.title = element_text(size = 8))

print(pdp_plot_glm_sub_MATMAP)

imp_plot_glm_sub_MATMAP$layers[[3]] <- NULL
ggarrange(pdp_plot_glm_sub_MATMAP, imp_plot_glm_sub_MATMAP,
          labels = "AUTO", ncol = 2, common.legend = F,
          legend = "bottom", widths = c(1.5, 0.5), hjust = -0.05, vjust = 1)

ggsave(here::here("plots/plots_pdp/pdp_plot_glm_subclades_MAT_MAP_litt_vars.png"),
       width = width_mm, height = width_mm * 0.5, units = "mm", dpi = 300, bg = "white")

# --- Low-abundance subclades ---

subs_lowAb_glm_MATMAP <- intersect(
  correct_order[-c(1, 4, 5, 12, 16, 17, 23, 25, 29)],
  unique(all_imp_sig_glm_sub_MATMAP$Subclade[
    !all_imp_sig_glm_sub_MATMAP$Subclade %in% c("IC", "IIG", "IIIA", "IVB")])
)

imp_plot_glm_sub_MATMAP_lowAb <- all_imp_sig_glm_sub_MATMAP %>%
  dplyr::filter(!Subclade %in% c("IC", "IIG", "IIIA", "IVB")) |>
  mutate(Subclade = factor(Subclade, levels = subs_lowAb_glm_MATMAP)) |>
  ggplot(aes(x = importance, y = feature, color = feature)) +
  scale_color_manual(values = allvars_pallete[feature_levels_MATMAP],
                     guide = guide_legend(ncol = 2)) +
  scale_y_discrete(limits = rev(feature_levels_MATMAP)) +
  scale_x_continuous(labels = scales::percent_format()) +
  ggh4x::facet_grid2(Subclade ~ ., scales = "free_y", space = "free_y",
                     strip = ggh4x::strip_themed(
                       background_y = ggh4x::elem_list_rect(
                         fill = subclade_colors_bp[subs_lowAb_glm_MATMAP]
                       )
                     )) +
  geom_point(position = position_dodge(width = 0.6), size = 3) +
  geom_errorbarh(aes(xmin = importance.05, xmax = importance.95),
                 position = position_dodge(width = 0.6), height = 0, linewidth = 1.5) +
  xlab("importance (RMSE)") +
  apatheme +
  theme(axis.text.y = element_blank(), axis.title.y = element_blank(),
        axis.line.y = element_blank(), strip.text = element_text(size = 10),
        panel.grid.major.x = element_line(colour = "grey90"),
        panel.border = element_rect(colour = "grey90", fill = NA, linewidth = 0.5),
        legend.spacing = unit(0.1, "cm"), legend.key.size = unit(0.3, "cm"),
        legend.margin = ggplot2::margin(0, 0, 0, 0),
        legend.text = element_text(size = 9))

pdp_plot_glm_sub_MATMAP_lowAb <- all_pdp_ci_glm_sub_MATMAP %>%
  mutate(Clade = case_when(Subclade %in% clade_I_subclades   ~ "I",
                           Subclade %in% clade_II_subclades  ~ "II",
                           Subclade %in% clade_III_subclades ~ "III",
                           Subclade %in% clade_IV_subclades  ~ "IV",
                           Subclade %in% clade_V_subclades   ~ "V")) %>%
  mutate(var_type = case_when(feature %in% clim_vars_MATMAP ~ "Climate",
                              feature %in% litter_vars       ~ "Litter chemistry",
                              TRUE ~ NA_character_)) |>
  mutate(var_type = factor(var_type, levels = c("Climate", "Litter chemistry")),
         feature  = gsub("_", " ", feature),
         feature  = factor(feature, levels = feature_levels_MATMAP)) |>
  dplyr::filter(!Subclade %in% c("IC", "IIG", "IIIA", "IVB")) |>
  ggplot(aes(x = value, y = mean_pred, color = Subclade, linetype = significant)) +
  geom_line(size = 1) +
  scale_y_continuous(labels = percent_format(scale = 100)) +
  scale_color_manual(values = subclade_colors_bp) +
  scale_linetype_manual(name = NULL,
                        values = c("TRUE" = "solid", "FALSE" = "longdash"),
                        labels = c(expression(P > 0.05), expression(P < 0.05))) +
  geom_rug(data = rug_glm_sub_MATMAP, aes(x = value), sides = "b", inherit.aes = FALSE) +
  ggh4x::facet_nested(Clade ~ var_type + feature, scales = "free_x") +
  labs(x = "Variable value", y = "Predicted relative abundance") +
  guides(color = "none", fill = "none",
         linetype = guide_legend(order = 1, override.aes = list(size = 1))) +
  apatheme +
  theme(axis.title.y = element_text(size = 8, margin = ggplot2::margin(r = 10)),
        axis.text.y = element_text(size = 8),
        strip.background = element_rect(fill = NA, colour = "black"),
        strip.text = element_text(size = 10),
        panel.grid.major.y = element_line(colour = "grey90"),
        text = element_text(size = 10),
        legend.position = c(0.02, 0.98), legend.justification = c(0, 1),
        legend.background = element_rect(fill = alpha("white", 0.7), colour = "black"),
        legend.key.size = unit(0.3, "cm"), legend.text = element_text(size = 6),
        legend.key.width = unit(1, "cm"), legend.key.height = unit(0.5, "cm"),
        legend.title = element_text(size = 8))

print(pdp_plot_glm_sub_MATMAP_lowAb)

ggarrange(pdp_plot_glm_sub_MATMAP_lowAb, imp_plot_glm_sub_MATMAP_lowAb,
          labels = "AUTO", ncol = 2, common.legend = F,
          legend = "bottom", widths = c(1.5, 0.5), hjust = -0.05, vjust = 1)

ggsave(here::here("plots/plots_pdp/pdp_plot_glm_subclades_MAT_MAP_litt_vars_lowAb.png"),
       width = width_mm, height = width_mm * 0.75, units = "mm", dpi = 300)
```

##### 2D heatmap PDPs — top 4 subclades, MAT & MAP (GLM)

```{r pdp-subclades-glm-heatmap-mat-map, message=FALSE, warning=FALSE, results='hold', echo=FALSE, fig.width=20, fig.height=12}
subclades_2d_pdp_glm_MATMAP <- cached(here::here("cache/subclades_2d_pdp_glm_MATMAP.rds"), {
  map_dfr(top4_subclades, function(sub) {
    message("Subclade: ", sub)
    mod  <- subclade_glm_results_MATMAP[[sub]]$glm_full
    X    <- subclade_glm_results_MATMAP[[sub]]$rug
    y    <- subclade_glm_results_MATMAP[[sub]]$glm_full$y
    pred <- Predictor$new(mod, data = X, y = y)

    map_dfr(clim_vars_MATMAP, function(cv) {
      map_dfr(litter_vars, function(lv) {
        message("  ", cv, " x ", lv)
        eff <- FeatureEffect$new(pred, feature = c(cv, lv), method = "pdp", grid.size = 20)
        eff$results |>
          dplyr::rename(clim_value = all_of(cv), litter_value = all_of(lv)) |>
          dplyr::mutate(clim_var = cv, litter_var = lv, Subclade = sub)
      })
    })
  })
})

heatmap_2d_glm_MATMAP <- function(sub) {
  subclades_2d_pdp_glm_MATMAP |>
    dplyr::filter(Subclade == sub) |>
    mutate(litter_var = gsub("_", " ", litter_var),
           litter_var = factor(litter_var, levels = gsub("_", " ", litter_vars)),
           clim_var   = factor(clim_var,   levels = clim_vars_MATMAP)) |>
    ggplot(aes(x = litter_value, y = clim_value, fill = .value)) +
    geom_raster(interpolate = TRUE) +
    scale_fill_viridis_c(name = "Predicted\nabundance",
                         labels = scales::percent_format(scale = 100)) +
    facet_grid(clim_var ~ litter_var, scales = "free") +
    labs(x = "Litter chemistry variable value",
         y = "Climate variable value", title = sub) +
    apatheme +
    theme(plot.title        = element_text(hjust = 0.5, face = "bold", size = 11,
                                           colour = subclade_colors_bp[sub]),
          strip.background  = element_rect(fill = NA, colour = "black"),
          strip.text        = element_text(size = 8),
          axis.text         = element_text(size = 7),
          axis.title        = element_text(size = 9),
          panel.spacing     = unit(0.3, "cm"),
          legend.key.height = unit(1, "cm"))
}

ggarrange(plotlist = lapply(top4_subclades, heatmap_2d_glm_MATMAP),
          ncol = 2, nrow = 2, common.legend = TRUE, legend = "right")

ggsave(here::here("plots/plots_pdp/pdp_heatmap_glm_top4_subclades_MATMAP.png"),
       width = 300, height = 200, units = "mm", dpi = 300, bg = "white")
```

##### 2D heatmap PDPs — significant predictors only, MAT & MAP (GLM)

```{r pdp-subclades-glm-heatmap-mat-map-sig, message=FALSE, warning=FALSE, results='hold', echo=FALSE, fig.width=16, fig.height=10}
sig_pairs_glm_MATMAP <- lapply(setNames(top4_subclades, top4_subclades), function(sub) {
  sig      <- subclade_glm_results_MATMAP[[sub]]$sig_vars_rfp
  sig_clim <- intersect(sig, clim_vars_MATMAP)
  sig_litt <- intersect(sig, litter_vars)
  expand.grid(clim_var = sig_clim, litter_var = sig_litt, stringsAsFactors = FALSE)
})
sig_pairs_glm_MATMAP <- Filter(function(x) nrow(x) > 0, sig_pairs_glm_MATMAP)

message("Subclades with significant MAT/MAP × litter pairs (GLM): ",
        paste(names(sig_pairs_glm_MATMAP), collapse = ", "))

subclades_2d_pdp_glm_MATMAP_sig <- cached(here::here("cache/subclades_2d_pdp_glm_MATMAP_sig.rds"), {
  map_dfr(names(sig_pairs_glm_MATMAP), function(sub) {
    message("Subclade: ", sub)
    mod  <- subclade_glm_results_MATMAP[[sub]]$glm_full
    X    <- subclade_glm_results_MATMAP[[sub]]$rug
    y    <- subclade_glm_results_MATMAP[[sub]]$glm_full$y
    pred <- Predictor$new(mod, data = X, y = y)

    pmap_dfr(sig_pairs_glm_MATMAP[[sub]], function(clim_var, litter_var) {
      message("  ", clim_var, " x ", litter_var)
      eff <- FeatureEffect$new(pred, feature = c(clim_var, litter_var),
                               method = "pdp", grid.size = 20)
      eff$results |>
        dplyr::rename(clim_value = all_of(clim_var), litter_value = all_of(litter_var)) |>
        dplyr::mutate(clim_var = clim_var, litter_var = litter_var, Subclade = sub)
    })
  })
})

heatmap_2d_glm_MATMAP_sig <- function(sub) {
  subclades_2d_pdp_glm_MATMAP_sig |>
    dplyr::filter(Subclade == sub) |>
    mutate(litter_var = gsub("_", " ", litter_var),
           litter_var = factor(litter_var, levels = gsub("_", " ", litter_vars)),
           clim_var   = factor(clim_var,   levels = clim_vars_MATMAP)) |>
    ggplot(aes(x = litter_value, y = clim_value, fill = .value)) +
    geom_raster(interpolate = TRUE) +
    scale_fill_viridis_c(name = "Predicted\nabundance",
                         labels = scales::percent_format(scale = 100),
                         limits = c(0, 0.60), oob = scales::squish) +
    facet_grid(clim_var ~ litter_var, scales = "free") +
    labs(x = "Litter chemistry variable value",
         y = "Climate variable value", title = sub) +
    apatheme +
    theme(plot.title        = element_text(hjust = 0.5, face = "bold", size = 11,
                                           colour = subclade_colors_bp[sub]),
          strip.background  = element_rect(fill = NA, colour = "black"),
          strip.text        = element_text(size = 8),
          axis.text         = element_text(size = 7),
          axis.title        = element_text(size = 9),
          panel.spacing     = unit(0.3, "cm"),
          legend.key.height = unit(1, "cm"))
}

pdp_heatmaps_glm_MATMAP_sig <- lapply(names(sig_pairs_glm_MATMAP), heatmap_2d_glm_MATMAP_sig)

ggarrange(plotlist = pdp_heatmaps_glm_MATMAP_sig,
          ncol = length(pdp_heatmaps_glm_MATMAP_sig), nrow = 1,
          common.legend = TRUE, legend = "right")

ggsave(here::here("plots/plots_pdp/pdp_heatmap_glm_top4_subclades_MATMAP_sig_vars.png"),
       width = 100 * length(pdp_heatmaps_glm_MATMAP_sig), height = 160,
       units = "mm", dpi = 300, bg = "white")
```
